---
title: "Figure_4_S7-8"
author: "Alexander Okamoto"
date: "2025-04-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r}
#load packages
library(readxl)
library(GenomicRanges)
library(tidyverse)
library(cowplot)
library(ggpubr)
library(regioneR) #for reading bed files as GRanges
library(ggnewscale)

#load libraries for TF analysis
library("monaLisa")
library(TFBSTools)
library(JASPAR2020)
library(BSgenome.Hsapiens.UCSC.hg38)
library(SummarizedExperiment)
library(mia)

#load custom functions for analysis
source("~/Desktop/Capellini_Lab/Weekly_Coding/RNA_Analysis_Functions.R") 
```

#get human/chimp/gorilla variants based on GAGDP and 1KGP datasets

```{bash, eval = F}
bedtools intersect -a  ~/Autopod_Data/corrected_multiinter_autopod_simple_hg18.bed -b  /home/aokamoto/GREATAPE/Pan_troglodytes_all_variants_MAF_005.vcf.gz -wa -c > ~/Autopod_Data/corrected_multiinter_autopod_chimp_vars.bed

bedtools intersect -a  ~/Autopod_Data/corrected_multiinter_autopod_simple_hg18.bed -b  /home/aokamoto/GREATAPE/Gorilla_all_variants_MAF_005.vcf.gz -wa -c > ~/Autopod_Data/corrected_multiinter_autopod_gorilla_vars.bed

bedtools intersect -a  ~/Autopod_Data/corrected_multiinter_autopod_simple_hg38.bed -b  /home/aokamoto/GREATAPE/Gorilla_all_variants_MAF_005.vcf.gz -wa -c > ~/Autopod_Data/corrected_multiinter_autopod_gorilla_vars.bed

#iterate through chromosomes 
for chr in {1..22} X Y
do

bedtools intersect -a  ~/Autopod_Data/corrected_multiinter_autopod_simple_hg38.bed -b  ~/1KG3/ALL.chr${chr}_MAF_005.vcf.gz -wa -c > ~/Autopod_Data/human_temp_chr${chr}.bed

done

do
name=$(echo $bed | cut -d'/' -f5 | cut -d'.' -f1)
echo $name

#reformat chromosome column to match
sed -i 's/chr//' ~/Autopod_Data/corrected_multiinter_autopod_simple_hg38.bed

#iterate through chromosomes 
for chr in {1..22} X Y
do

#subset bed file by chromosome
awk -F'\t' -v awkvar=$chr '$1 == awkvar' ~/Autopod_Data/corrected_multiinter_autopod_simple_hg38.bed > temp_chr${chr}.bed
#do bedtools intersection to count SNPs per region
seq 200 | parallel "bedtools intersect -a temp_chr${chr}.bed -b ~/1KG3/Human_Reps/Human_subset{}_chr${chr}.vcf -wa -c > human_vals_temp{}_${chr}.txt"
#re
rm temp_chr*.bed
done

cat human_vals_temp*.txt | sort -k 1,1 -k2,2n > ~/Autopod_Data/corrected_multiinter_autopod_human_vars.bed

#subset bed file by chromosome
awk -F'\t' -v awkvar=$chr '$1 == awkvar' $bed > temp_chr${chr}_${name}.bed
#do bedtools intersection to count SNPs per region
seq 200 | parallel "bedtools intersect -a temp_chr${chr}_${name}.bed -b ~/1KG3/Human_Reps/Human_subset{}_chr${chr}.vcf -c > human_${name}_vals_temp{}_${chr}.txt"
#re
rm temp_chr*.bed
done
```

#prepare for structural variant analysis

```{r}
inversions <- "~/Desktop/Capellini_Lab/Structural_Variant_Beds/human_specific_inversions_hg38.bed"
SDRs <- "~/Desktop/Capellini_Lab/Structural_Variant_Beds/human_SDRs_hg38.bed"

inversions_SDRs <- "~/Desktop/Capellini_Lab/Structural_Variant_Beds/human_inversions_SDRs_hg38.bed"

hg38_sizes <- "~/Desktop/Capellini_Lab/Capellini_Lab_Peak_Sets/Other_Peak_sets/hg38_clean_sizes.txt"

system(paste("cat", inversions, SDRs, "| sort -k 1,1 -k2,2n | bedtools merge -i stdin >", inversions_SDRs , sep = " "))

flank_size <- 500000
inversions_flank <- gsub(x = inversions, pattern = "_hg38.bed", replacement = paste("flank_", flank_size, "_hg38.bed", sep = ""))
SDRs_flank <- gsub(x = SDRs, pattern = "_hg38.bed", replacement = paste("flank_", flank_size, "_hg38.bed", sep = ""))

system(paste("bedtools slop -i", inversions, "-g", hg38_sizes, "-b", flank_size, "| bedtools subtract -a stdin -b", inversions, "| sort -k 1,1 -k2,2n | bedtools merge -i stdin >", inversions_flank  , sep = " "))

system(paste("bedtools slop -i", SDRs, "-g", hg38_sizes, "-b", flank_size, "| bedtools subtract -a stdin -b", SDRs, "| sort -k 1,1 -k2,2n | bedtools merge -i stdin >", SDRs_flank  , sep = " "))
```

## find/count regions with specific tissue properties

```{r}
#load files with merged ATAC-seq peak sets
#generated in Figure_2_RNA_ATAC_PCA.R

multiinter_both_df <- read.table( "~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod.bed")
colnames(multiinter_both_df) <- c("chr", "start", "end", "num", "list", paste(c("FM1", "FM2", "FM3", "FM4", "FM5", "FP1", "FP2", "FP3", "FP4", "FP5", "HM1", "HM2", "HM3", "HM4", "HM5", "HP1", "HP2", "HP3", "HP4", "HP5"), "_Early", sep = ""), paste(c("FM1", "FM2", "FM3", "FM4", "FM5", "FP1", "FP2", "FP3", "FP4", "FP5", "HM1", "HM2", "HM3", "HM4", "HM5", "HP1", "HP2", "HP3", "HP4", "HP5"), "_Late", sep = ""))
multiinter_both_df$ID <- paste(multiinter_both_df$chr, multiinter_both_df$start, multiinter_both_df$end, sep = "_")

#used for converting to mouse coordinates
# write_bed(multiinter_both_df %>% dplyr::select(chr, start, end, ID), ncol = 4, filename = "~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_simple_hg38.bed")
# liftover_to_hg38(bedfile = "~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_simple_hg38.bed", chain = "hg38tohg19")
# 
# liftover_to_hg38(bedfile = "~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_simple_hg19.bed", chain = "hg19tohg18")

#var files contain number of common variants (MAF > 0.05)
#load and merge var files for human, chimp, gorilla

multiinter_df_human <- read.delim(file = "~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_human_vars.bed", header = F, sep = "\t")
names(multiinter_df_human) <- c("chr", "start", "end", "ID", "human_SNPs")

multiinter_df_human2 <- multiinter_df_human %>% dplyr::select(ID, human_SNPs)

#needed to cleanup up output since human variants per chromosome had to be split due avoid errors
# names(multiinter_df_human_v1) <- c("chr", "start", "end", "ID", "human_SNPs_raw")
# multiinter_df_human <- multiinter_df_human_v1 %>% group_by(ID) %>% mutate(human_SNPs = max(human_SNPs_raw)) %>% dplyr::select(-human_SNPs_raw) %>% unique()
# write_bed(filename = "~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_human_vars.bed", data = multiinter_df_human, ncol = 5)

multiinter_df_chimp <- read.delim(file = "~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_chimp_vars.bed", header = F, sep = "\t")
names(multiinter_df_chimp) <- c("chr", "start", "end", "ID", "chimp_SNPs")
multiinter_df_chimp2 <- multiinter_df_chimp %>% dplyr::select(ID, chimp_SNPs)

multiinter_df_gorilla <- read.delim(file = "~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_gorilla_vars.bed", header = F, sep = "\t")
names(multiinter_df_gorilla) <- c("chr", "start", "end", "ID", "gorilla_SNPs")

multiinter_df_gorilla2 <- multiinter_df_gorilla %>% dplyr::select(ID, gorilla_SNPs)

multiinter_df_ape <- merge(x = multiinter_df_chimp2 %>% dplyr::select(ID, chimp_SNPs), y = multiinter_df_gorilla2 %>% dplyr::select(ID, gorilla_SNPs), all.x =T, all.y = T, by ="ID") %>% merge(y = multiinter_df_human2, all.x =T, all.y = T, by ="ID")

#get subset of regions overlapping non-autopod tissues
#All_NOT_AUTOPOD_hg38.bed contains merged peaks from long bones, girdles, and vertebral column from previous studies (see Methods)

system("bedtools intersect -a ~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod.bed -b ~/Desktop/Capellini_Lab/Weekly_Coding/Capellini_Beds/All_NOT_AUTOPOD_hg38.bed -wa | uniq > ~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_shared_non_autopod.bed")

multiinter_df_non_autopod <- read.delim(file = "~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_shared_non_autopod.bed", header = F, sep = "\t")
colnames(multiinter_df_non_autopod) <- c("chr", "start", "end", "num", "list", paste(c("FM1", "FM2", "FM3", "FM4", "FM5", "FP1", "FP2", "FP3", "FP4", "FP5", "HM1", "HM2", "HM3", "HM4", "HM5", "HP1", "HP2", "HP3", "HP4", "HP5"), "E", sep = ""), paste(c("FM1", "FM2", "FM3", "FM4", "FM5", "FP1", "FP2", "FP3", "FP4", "FP5", "HM1", "HM2", "HM3", "HM4", "HM5", "HP1", "HP2", "HP3", "HP4", "HP5"), "L", sep = ""))
multiinter_df_non_autopod$ID <- paste(multiinter_df_non_autopod$chr, multiinter_df_non_autopod$start, multiinter_df_non_autopod$end, sep = "_")

multiinter_df_non_autopod <-merge(x = multiinter_df_non_autopod, y = multiinter_df_chimp, all.x =T, by = "ID")

multiinter_autopod_only <- multiinter_both_df %>% dplyr::filter(!ID %in% multiinter_df_non_autopod$ID) 

#multiinter_df_non_autopod <- read.delim(file = "~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_shared_non_autopod.bed", header = F, sep = "\t")
colnames(multiinter_df_non_autopod) <- c("chr", "start", "end", "num", "list", paste(c("FM1", "FM2", "FM3", "FM4", "FM5", "FP1", "FP2", "FP3", "FP4", "FP5", "HM1", "HM2", "HM3", "HM4", "HM5", "HP1", "HP2", "HP3", "HP4", "HP5"), "E", sep = ""), paste(c("FM1", "FM2", "FM3", "FM4", "FM5", "FP1", "FP2", "FP3", "FP4", "FP5", "HM1", "HM2", "HM3", "HM4", "HM5", "HP1", "HP2", "HP3", "HP4", "HP5"), "L", sep = ""))

multiinter_HARs <- read.delim(file = "~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_HARs.bed", header = F, sep = "\t")

multiinter_HARs_IDs <- paste(multiinter_HARs$V1, multiinter_HARs$V2, multiinter_HARs$V3, sep = "_")

#hCONDELs next
multiinter_hCONDELs <- read.delim(file = "~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_hCONDEL.bed", header = F, sep = "\t")

multiinter_hCONDEL_IDs <- paste(multiinter_hCONDELs$V1, multiinter_hCONDELs$V2, multiinter_hCONDELs$V3, sep = "_")

#now new T2T HAQERs
system("bedtools intersect -a ~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod.bed -b ~/Desktop/Capellini_Lab/Structural_Variant_Beds/HAQERS_T2T_hg38.bed -wa -wb | uniq > ~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_T2T_HAQERs.bed")

multiinter_HAQERs <- read.delim(file = "~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_T2T_HAQERs.bed", header = F, sep = "\t")

multiinter_HAQERs_IDs <- paste(multiinter_HAQERs$V1, multiinter_HAQERs$V2, multiinter_HAQERs$V3, sep = "_")

#now mouse conservation

system("bedtools intersect -a ~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod.bed -b ~/Desktop/Capellini_Lab/Mouse_ATAC/mouse_multiinter_simple_hg38.bed -wa -wb | uniq > ~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_mouse.bed")

multiinter_mouse <- read.delim(file = "~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_mouse.bed", header = F, sep = "\t")
colnames(multiinter_mouse) <- c("chr", "start", "end", "num", "list", paste(c("FM1", "FM2", "FM3", "FM4", "FM5", "FP1", "FP2", "FP3", "FP4", "FP5", "HM1", "HM2", "HM3", "HM4", "HM5", "HP1", "HP2", "HP3", "HP4", "HP5"), "E", sep = ""), paste(c("FM1", "FM2", "FM3", "FM4", "FM5", "FP1", "FP2", "FP3", "FP4", "FP5", "HM1", "HM2", "HM3", "HM4", "HM5", "HP1", "HP2", "HP3", "HP4", "HP5"), "L", sep = ""), "mus_chr", "mus_start", "mus_end", "mus_list", "mus_list2")
multiinter_mouse_IDs <- paste(multiinter_mouse$chr, multiinter_mouse$start, multiinter_mouse$end, sep = "_")

#structural variant analysis

system("bedtools intersect -a ~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod.bed -b ~/Desktop/Capellini_Lab/Structural_Variant_Beds/human_SDRs_hg38.bed -wa -wb | uniq > ~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_SDRs.bed")

multiinter_SDRs <- read.delim(file = "~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_SDRs.bed", header = F, sep = "\t")

multiinter_SDRs_IDs <- paste(multiinter_SDRs$V1, multiinter_SDRs$V2, multiinter_SDRs$V3, sep = "_")

system("bedtools intersect -a ~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod.bed -b ~/Desktop/Capellini_Lab/Structural_Variant_Beds/human_SDRsflank_500000_hg38.bed -wa -wb | uniq > ~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_SDRs_flank1.bed")

multiinter_SDRs_flank1 <- read.delim(file = "~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_SDRs_flank1.bed", header = F, sep = "\t")

multiinter_SDRs_flank1_IDs <- paste(multiinter_SDRs_flank1$V1, multiinter_SDRs_flank1$V2, multiinter_SDRs_flank1$V3, sep = "_")

system("bedtools intersect -a ~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod.bed -b ~/Desktop/Capellini_Lab/Structural_Variant_Beds/human_SDRsflank_1000000_hg38.bed -wa -wb | uniq > ~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_SDRs_flank2.bed")

multiinter_SDRs_flank2 <- read.delim(file = "~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_SDRs_flank2.bed", header = F, sep = "\t")

multiinter_SDRs_flank2_IDs <- paste(multiinter_SDRs_flank2$V1, multiinter_SDRs_flank2$V2, multiinter_SDRs_flank2$V3, sep = "_")

system("bedtools intersect -a ~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod.bed -b ~/Desktop/Capellini_Lab/Structural_Variant_Beds/human_specific_inversions_hg38.bed -wa -wb | uniq > ~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_inversions.bed")

multiinter_inversions <- read.delim(file = "~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_inversions.bed", header = F, sep = "\t")

multiinter_inversions_IDs <- paste(multiinter_inversions$V1, multiinter_inversions$V2, multiinter_inversions$V3, sep = "_")

system("bedtools intersect -a ~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod.bed -b ~/Desktop/Capellini_Lab/Structural_Variant_Beds/human_specific_inversionsflank_500000_hg38.bed -wa -wb | uniq > ~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_inversions_flank1.bed")

multiinter_inversions_flank1 <- read.delim(file = "~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_inversions_flank1.bed", header = F, sep = "\t")

multiinter_inversions_flank1_IDs <- paste(multiinter_inversions_flank1$V1, multiinter_inversions_flank1$V2, multiinter_inversions_flank1$V3, sep = "_")

system("bedtools intersect -a ~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod.bed -b ~/Desktop/Capellini_Lab/Structural_Variant_Beds/human_specific_inversionsflank_1000000_hg38.bed -wa -wb | uniq > ~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_inversions_flank2.bed")

multiinter_inversions_flank2 <- read.delim(file = "~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_inversions_flank2.bed", header = F, sep = "\t")

multiinter_inversions_flank2_IDs <- paste(multiinter_inversions_flank2$V1, multiinter_inversions_flank2$V2, multiinter_inversions_flank2$V3, sep = "_")
```

#create figure to visualize results
#create functions to do parts of analysis

```{r}
#create smaller function to run the Fisher's exact tests
pairwise_fisher_test <- function(val1, val2, val3, cat_list, df){
    print(paste(cat_list[val1], "VS", cat_list[val2]))
    print("HARs")
    table12_HAR <- table(df$HAR[which(!df$category %in% cat_list[val3])], df$category[which(!df$category %in% cat_list[val3])])
  print(fisher.test(table12_HAR))
  print("hCONDELs")
    table12_hCONDEL <- table(df$hCONDEL[which(!df$category %in% cat_list[val3])], df$category[which(!df$category %in% cat_list[val3])])
  print(fisher.test(table12_hCONDEL))
  print("SDRs")
    table12_SDR <- table(df$SDR[which(!df$category %in% cat_list[val3])], df$category[which(!df$category %in% cat_list[val3])])
  print(fisher.test(table12_SDR))
  print("Inversion")
  table12_inversion <- table(df$inversion[which(!df$category %in% cat_list[val3])], df$category[which(!df$category %in% cat_list[val3])])
  print(fisher.test(table12_inversion))
}

#takes a df with two columns, first is the chr_start_end of the autopod only multint data, column 2 is the categorical data desired

divide_autopod_only_data_plotting <- function(divisor_df, setID, exclude_SDR = F){
  
  #prepare dataframe to use for determining overlaps
  divisor_df2 <- divisor_df 
  colnames(divisor_df2) <- c("ID", "category")
  divisor_df2 <- divisor_df2 %>% merge(y = multiinter_df_ape, all.x =T, by = "ID")
  categories <- unique(divisor_df2$category)
  
    #create dataframe for results
  results_df <- data.frame(overlap = rep(x = c("HAR", "HAQER", "hCONDEL", "SDR", "SDR_Flank_500k", "SDR_Flank_1M", "Inversion", "Inversion_Flank_500k", "Inversion_Flank_1M", "human_SNPs","chimp_SNPs", "gorilla_SNPs"), each = length(categories)), 
                           set = setID, 
                           category = categories, 
                           value = NA, 
                           set_total = NA,
                           norm_value = NA)
  #skip structurally divergent regions for mouse conserved analysis
  if(exclude_SDR == T){
     results_df  <- results_df %>% dplyr::filter(! overlap %in% c("SDR", "SDR_Flank_500k", "SDR_Flank_1M"))
  }
  
  #add HAR/hCONDEL data
  divisor_df2$HAR <- "none"
  divisor_df2$HAR[which(divisor_df2$ID %in% multiinter_HARs_IDs)] <- "HAR"
  divisor_df2$hCONDEL <- "none"
  divisor_df2$hCONDEL[which(divisor_df2$ID %in% 
                              multiinter_hCONDEL_IDs)] <- "hCONDEL"
   divisor_df2$HAQER <- "none"
  divisor_df2$HAQER[which(divisor_df2$ID %in% 
                              multiinter_HAQERs_IDs)] <- "HAQER"
  divisor_df2$SDR <- "none"
  divisor_df2$SDR[which(divisor_df2$ID %in% 
                              multiinter_SDRs_IDs)] <- "SDR"
  divisor_df2$SDR_flank1 <- "none"
  divisor_df2$SDR_flank1[which(divisor_df2$ID %in% 
                              multiinter_SDRs_flank1_IDs)] <- "SDR"
  divisor_df2$SDR_flank2 <- "none"
  divisor_df2$SDR_flank2[which(divisor_df2$ID %in% 
                              multiinter_SDRs_flank2_IDs)] <- "SDR"
  divisor_df2$inversion <- "none"
  divisor_df2$inversion[which(divisor_df2$ID %in% 
                              multiinter_inversions_IDs)] <- "inversion"
  divisor_df2$inversion_flank1 <- "none"
  divisor_df2$inversion_flank1[which(divisor_df2$ID %in% 
                              multiinter_inversions_flank1_IDs)] <- "inversion"
  divisor_df2$inversion_flank2 <- "none"
  divisor_df2$inversion_flank2[which(divisor_df2$ID %in% 
                              multiinter_inversions_flank2_IDs)] <- "inversion"

  for (i in 1:length(categories)){
    results_df$set_total[which(results_df$category == categories[i])] <- divisor_df2 %>%  
      dplyr::filter(category == categories[i]) %>% nrow()
    results_df$value[which((results_df$category == categories[i]) & (results_df$overlap == "HAR"))] <- divisor_df2 %>%  
      dplyr::filter(category == categories[i], HAR == "HAR") %>% nrow()
    results_df$value[which((results_df$category == categories[i]) & (results_df$overlap == "hCONDEL"))] <- divisor_df2 %>%  
      dplyr::filter(category == categories[i], hCONDEL == "hCONDEL") %>% nrow()
    results_df$value[which((results_df$category == categories[i]) & (results_df$overlap == "HAQER"))] <- divisor_df2 %>%  
      dplyr::filter(category == categories[i], HAQER == "HAQER") %>% nrow()
    #check if SDR overlaps should be excluded 
    if(exclude_SDR == F){
    results_df$value[which((results_df$category == categories[i]) & (results_df$overlap == "SDR"))] <- divisor_df2 %>%  
      dplyr::filter(category == categories[i], SDR == "SDR") %>% nrow()
    results_df$value[which((results_df$category == categories[i]) & (results_df$overlap == "SDR_Flank_500k"))] <- divisor_df2 %>%  
      dplyr::filter(category == categories[i], SDR_flank1 == "SDR") %>% nrow()
    results_df$value[which((results_df$category == categories[i]) & (results_df$overlap == "SDR_Flank_1M"))] <- divisor_df2 %>%  
      dplyr::filter(category == categories[i], SDR_flank2 == "SDR") %>% nrow()
    }
    results_df$value[which((results_df$category == categories[i]) & (results_df$overlap == "Inversion"))] <- divisor_df2 %>%  
      dplyr::filter(category == categories[i], inversion == "inversion") %>% nrow()
    results_df$value[which((results_df$category == categories[i]) & (results_df$overlap == "Inversion_Flank_500k"))]<- divisor_df2 %>%  
      dplyr::filter(category == categories[i], inversion_flank1 == "inversion") %>% nrow()
    results_df$value[which((results_df$category == categories[i]) & (results_df$overlap == "Inversion_Flank_1M"))] <- divisor_df2 %>%  
      dplyr::filter(category == categories[i], inversion_flank2 == "inversion") %>% nrow()
    #add the intraspecific diversity data
    results_df$value[which((results_df$category == categories[i]) & (results_df$overlap == "human_SNPs"))] <- divisor_df2 %>%  
      dplyr::filter(category == categories[i]) %>% pull(human_SNPs) %>% sum(na.rm=TRUE)
    results_df$set_total[which((results_df$category == categories[i]) & (results_df$overlap == "human_SNPs"))] <- divisor_df2 %>%  
      dplyr::filter(category == categories[i]) %>% 
      dplyr::filter(!is.na(human_SNPs)) %>% 
      nrow()
    #add chimp
    results_df$value[which((results_df$category == categories[i]) & (results_df$overlap == "chimp_SNPs"))] <- divisor_df2 %>%  
      dplyr::filter(category == categories[i]) %>% pull(chimp_SNPs) %>% sum(na.rm=TRUE)
    results_df$set_total[which((results_df$category == categories[i]) & (results_df$overlap == "chimp_SNPs"))] <- divisor_df2 %>%  
      dplyr::filter(category == categories[i]) %>% 
      dplyr::filter(!is.na(chimp_SNPs)) %>% 
      nrow()
    #add gorilla
    results_df$value[which((results_df$category == categories[i]) & (results_df$overlap == "gorilla_SNPs"))] <- divisor_df2 %>%  
      dplyr::filter(category == categories[i]) %>% pull(gorilla_SNPs) %>% sum(na.rm=TRUE)
    results_df$set_total[which((results_df$category == categories[i]) & (results_df$overlap == "gorilla_SNPs"))] <- divisor_df2 %>%  
      dplyr::filter(category == categories[i]) %>% 
      dplyr::filter(!is.na(gorilla_SNPs)) %>% 
      nrow()
    
  } #end loop

  results_df$norm_value <- results_df$value/results_df$set_total
  return(results_df)
} #end function

#takes a df with two columns, first is the chr_start_end of the autopod only multint data, column 2 is the categorical data desired

divide_autopod_only_stat_plotting <- function(divisor_df, setID, exclude_SDR = F){
  
  #prepare dataframe to use for determining overlaps
  divisor_df2 <- divisor_df 
  colnames(divisor_df2) <- c("ID", "category")
  divisor_df2 <- divisor_df2 %>% merge(y = multiinter_df_ape, all.x =T, by = "ID")
  #fix order of categories to avoid an error
  categories <- unique(divisor_df2$category)
  if(categories[1] %in% c("FOOT", "SHARED", "HAND")){
    categories <- c("FOOT", "HAND", "SHARED")
  }
  if(categories[1] %in% c("PHALANGEAL", "METAPODIAL", "BOTH_REGIONS")){
    categories <- c("PHALANGEAL", "METAPODIAL", "BOTH_REGIONS")
  }
    if(categories[1] %in% c("EARLY", "LATE", "BOTH_STAGES")){
    categories <- c("EARLY", "LATE", "BOTH_STAGES")
  }
  
    #create dataframe for results
  results_df <- data.frame(overlap = rep(x = c("HAR", "HAQER", "hCONDEL", "SDR", "SDR_Flank_500k", "SDR_Flank_1M", "Inversion", "Inversion_Flank_500k", "Inversion_Flank_1M", "human_SNPs", "chimp_SNPs", "gorilla_SNPs"), each = 3), 
                           comparison = NA, 
                           set = setID,
                           pval = NA, 
                           test = "exact"
                           )

  #add HAR/hCONDEL data
  divisor_df2$HAR <- "none"
  divisor_df2$HAR[which(divisor_df2$ID %in% multiinter_HARs_IDs)] <- "HAR"
  divisor_df2$hCONDEL <- "none"
  divisor_df2$hCONDEL[which(divisor_df2$ID %in% 
                              multiinter_hCONDEL_IDs)] <- "hCONDEL"
   divisor_df2$HAQER <- "none"
  divisor_df2$HAQER[which(divisor_df2$ID %in% 
                              multiinter_HAQERs_IDs)] <- "HAQER"
  divisor_df2$SDR <- "none"
  divisor_df2$SDR[which(divisor_df2$ID %in% 
                              multiinter_SDRs_IDs)] <- "SDR"
  divisor_df2$SDR_Flank_500k <- "none"
  divisor_df2$SDR_Flank_500k[which(divisor_df2$ID %in% 
                              multiinter_SDRs_flank1_IDs)] <- "SDR"
  divisor_df2$SDR_Flank_1M <- "none"
  divisor_df2$SDR_Flank_1M[which(divisor_df2$ID %in% 
                              multiinter_SDRs_flank2_IDs)] <- "SDR"
  divisor_df2$Inversion <- "none"
  divisor_df2$Inversion[which(divisor_df2$ID %in% 
                              multiinter_inversions_IDs)] <- "inversion"
  divisor_df2$Inversion_Flank_500k <- "none"
  divisor_df2$Inversion_Flank_500k[which(divisor_df2$ID %in% 
                              multiinter_inversions_flank1_IDs)] <- "inversion"
  divisor_df2$Inversion_Flank_1M <- "none"
  divisor_df2$Inversion_Flank_1M[which(divisor_df2$ID %in% 
                              multiinter_inversions_flank2_IDs)] <- "inversion"
  
   #run test for 
  if(length(categories) == 3){
    #add comparion values
    comparisons_list <- c(
      paste(categories[1], "_vs_", categories[2], sep = ""),
      paste(categories[1], "_vs_", categories[3], sep = ""),
      paste(categories[2], "_vs_", categories[3], sep = "")
      )
    
    results_df$comparison <- rep(comparisons_list, times = nrow(results_df)/3)
    
    #first Fisher's exact tests
    fisher_test <- c("HAR", "HAQER", "hCONDEL", "SDR", "SDR_Flank_500k", "SDR_Flank_1M", "Inversion", "Inversion_Flank_500k", "Inversion_Flank_1M")
   if(exclude_SDR == T) {
     fisher_test <- c("HAR", "HAQER", "hCONDEL", "SDR", "Inversion")
   }
    for (i in fisher_test){
      #get subset table
      table12 <- table(
        divisor_df2[,which(colnames(divisor_df2) == i)][which(!divisor_df2$category %in% categories[3])], 
        divisor_df2$category[which(!divisor_df2$category %in% categories[3])]
        )
    if(dim(table12)[1] == 1){
      results_df$pval[which((results_df$overlap == i) & (results_df$comparison == comparisons_list[1]))] <- NA
    } else{
      #perform Fisher's exact test
      test12 <- fisher.test(table12, workspace = 2e8)
      #add pvalue to results table
      results_df$pval[which((results_df$overlap == i) & (results_df$comparison == comparisons_list[1]))] <- test12$p.value
    }
      
      #get subset table
      table13 <- table(
        divisor_df2[,which(colnames(divisor_df2) == i)][which(!divisor_df2$category %in% categories[2])], 
        divisor_df2$category[which(!divisor_df2$category %in% categories[2])]
        )
    if(dim(table13)[1] == 1){
      results_df$pval[which((results_df$overlap == i) & (results_df$comparison == comparisons_list[2]))] <- NA
    } else{
      #perform Fisher's exact test
      test13 <- fisher.test(table13, workspace = 2e8)
      #add pvalue to results table
      results_df$pval[which((results_df$overlap == i) & (results_df$comparison == comparisons_list[2]))] <- test13$p.value
    }
      #get subset table
      table23 <- table(
        divisor_df2[,which(colnames(divisor_df2) == i)][which(!divisor_df2$category %in% categories[1])], 
        divisor_df2$category[which(!divisor_df2$category %in% categories[1])]
        )
          if(dim(table13)[1] == 1){
      results_df$pval[which((results_df$overlap == i) & (results_df$comparison == comparisons_list[3]))] <- NA
    } else{
      #perform Fisher's exact test
      test23 <- fisher.test(table23, workspace = 2e8)
      #add pvalue to results table
      results_df$pval[which((results_df$overlap == i) & (results_df$comparison == comparisons_list[3]))] <- test23$p.value
    }
      }
    
    #now for Wilcoxon test
     results_df$test[which(results_df$overlap %in% c("human_SNPs", "chimp_SNPs", "gorilla_SNPs"))] <- "wilcoxon"
    for (i in c("human_SNPs", "chimp_SNPs", "gorilla_SNPs")){
      divisor_df3 <-  divisor_df2[!is.na(divisor_df2[,which(colnames(divisor_df2) == i)]), ]
      #get subset table
      test12 <- wilcox.test(
        divisor_df3[,which(colnames(divisor_df3) == i)][which(divisor_df3$category == categories[1])], 
        divisor_df3[,which(colnames(divisor_df3) == i)][which(divisor_df3$category == categories[2])]
        )
      #add pvalue to results table
      results_df$pval[which((results_df$overlap == i) & (results_df$comparison == comparisons_list[1]))] <- test12$p.value
      #get subset table
      test13 <- wilcox.test(
        divisor_df3[,which(colnames(divisor_df3) == i)][which(divisor_df3$category == categories[1])], 
        divisor_df3[,which(colnames(divisor_df3) == i)][which(divisor_df3$category == categories[3])]
        )
      #add pvalue to results table
      results_df$pval[which((results_df$overlap == i) & (results_df$comparison == comparisons_list[2]))] <- test13$p.value
      #get subset table
      test23 <- wilcox.test(
        divisor_df3[,which(colnames(divisor_df3) == i)][which(divisor_df3$category == categories[2])], 
        divisor_df3[,which(colnames(divisor_df3) == i)][which(divisor_df3$category == categories[3])]
        )
      #add pvalue to results table
      results_df$pval[which((results_df$overlap == i) & (results_df$comparison == comparisons_list[3]))] <- test23$p.value
      
    } 
  }
  return(results_df)
} #end function
```

#repeat fixed human versus chimp SNP analysis to generate necessary file

```{r}

bedfile <- "~/Desktop/Capellini_Lab/Human_ATAC/corrected_multiinter_autopod_simple_hg38.bed"
  #gorilla requires generating a custom genome package
  #library(w)
  # forgeBSgenomeDataPkgFromUCSC(
  #     genome="gorGor6",
  #     organism="Gorilla gorilla",
  #     pkg_maintainer="Alexander Okamoto <aokamoto@g.harvard.edu>"
  # )
  #devtools::build("./BSgenome.Ggorilla.UCSC.gorGor6")
  #devtools::check_built("./BSgenome.Ggorilla.UCSC.gorGor6")
  #devtools::install_local("./BSgenome.Ggorilla.UCSC.gorGor6")
  
  #load packages
  
  require(tidyverse)
  require(BSgenome.Hsapiens.UCSC.hg38)
  require(BSgenome.Ptroglodytes.UCSC.panTro6)
  require(BSgenome.Ggorilla.UCSC.gorGor6)
  require(Biostrings)
  require(GenomicRanges)
  #bedfile <- "~/Desktop/Capellini_Lab/Human_ATAC/hand_only_both_tp.sorted_padded250.bed"
  #load genomes & chains
  chimp_genome <- BSgenome.Ptroglodytes.UCSC.panTro6
  chimp_chain <- "~/Desktop/Polymorphisms_Project/LiftOver/Chains/hg38ToPanTro6.over.chain"
  gorilla_genome <- BSgenome.Ggorilla.UCSC.gorGor6
  gorilla_chain <- "~/Desktop/Polymorphisms_Project/LiftOver/Chains/hg38.gorGor6.all.chain"
  
  #remove human & chimp SNP positions/repeat elements
  system(paste("bedtools subtract -a ", bedfile, " -b ~/Desktop/Capellini_Lab/Capellini_Lab_Peak_Sets/Other_Peak_sets/ENCODE_BLACKLIST_hg38.bed | ",
               "bedtools subtract -a stdin -b ~/Desktop/Capellini_Lab/Weekly_Coding/variants_dbSNP151_autopod_hg38.bed | bedtools subtract -a stdin -b ~/Desktop/Capellini_Lab/Weekly_Coding/greatape_autopod_SNPs_hg38.bed > ",
               "~/Desktop/TEMP_ATAC_DIR/temp_human_ape_bps.bed", 
               sep = ""
  ))
  
  #load bedfile and start processing
  bed_table <- read.delim("~/Desktop/TEMP_ATAC_DIR/temp_human_ape_bps.bed", sep = "\t", header = F)
  #add required column names for easier calling later on
  colnames(bed_table)[1:4] <- c("chr", "start", "end", "ID")
  bed_table <- bed_table %>% 
    dplyr::filter(chr %in% c(paste("chr", 1:22, sep = ""), "chrX", "chrY"))
  
  #get human DNA sequences
  bed_table$human_seqs <- bed_table %>% 
    dplyr::select(chr, start, end) %>% 
    dplyr::filter(chr %in% c(paste("chr", 1:22, sep = ""), "chrX", "chrY")) %>% 
    makeGRangesFromDataFrame() %>% 
    getSeq(x=BSgenome.Hsapiens.UCSC.hg38) %>% 
    as.character()
  #get lengths of sequence to expect
  bed_table$human_length <- bed_table$end - bed_table$start + 1
  #create stable ID value
  bed_table$ID2 <- paste(bed_table$chr, bed_table$start, bed_table$end, sep = "")
  
  #get ape DNA sequences
  
  #get granges of liftover object
  chimp_granges <- bed_table %>% 
    dplyr::filter(chr %in% c(paste("chr", 1:22, sep = ""), "chrX", "chrY")) %>% 
    makeGRangesFromDataFrame(keep.extra.columns = TRUE) %>% 
    liftOver(chain = import.chain(chimp_chain)) %>% 
    unlist()
  
  #make a dataframe to store extracted bits of sequence
  chimp_info <-  chimp_granges %>% as.data.frame()
  chimp_info$seq <- chimp_granges %>% getSeq(x = chimp_genome)  %>% 
    as.character()
  chimp_concat_info <- chimp_info %>% 
    dplyr::select(seq, ID2, width) %>% 
    group_by(ID2) %>% 
    mutate(chimp_seq = paste0(seq, collapse = ""), chimp_length = sum(width)) %>%
    dplyr::select(ID2, chimp_seq, chimp_length) %>% 
    unique()
  
  #get granges of liftover object
  gorilla_granges <- bed_table %>% 
    dplyr::filter(chr %in% c(paste("chr", 1:22, sep = ""), "chrX", "chrY")) %>% 
    makeGRangesFromDataFrame(keep.extra.columns = TRUE) %>% 
    liftOver(chain = import.chain(gorilla_chain)) %>% 
    unlist()
  
  #make a dataframe to store extracted bits of sequence
  gorilla_info <-  gorilla_granges %>% as.data.frame()
  gorilla_info$seq <- gorilla_granges %>% getSeq(x = gorilla_genome)  %>% 
    as.character()
  gorilla_concat_info <- gorilla_info %>% 
    dplyr::select(seq, ID2, width) %>% 
    group_by(ID2) %>% 
    mutate(gorilla_seq = paste0(seq, collapse = ""), gorilla_length = sum(width)) %>% 
    dplyr::select(ID2, gorilla_seq, gorilla_length) %>% 
    unique()
  
  #combined ape and human bedtables
  combined_table <- merge(bed_table, chimp_concat_info, by = 'ID2') %>% merge(gorilla_concat_info, by = 'ID2')
  #filter by same sequence length, remove any regions which match even less than expected by chance
  reduced_table <- combined_table  %>% 
    as_tibble() %>%
    dplyr::filter(human_length == chimp_length & human_length == gorilla_length) %>% 
    group_by(ID2) %>% 
    rowwise() %>% 
    mutate(bps_difs = compare_DNA_polarized(human_seqs, chimp_seq, gorilla_seq), chimp_bps_difs = compare_DNA_polarized(chimp_seq, human_seqs, gorilla_seq)) 
  #%>% 
  #  mutate(per_bp_difs = bps_difs/human_length) %>% dplyr::filter(per_bp_difs < 0.75)
  reduced_table$per_bp_difs <- reduced_table$bps_difs/reduced_table$human_length
  reduced_table$chimp_per_bp_difs <- reduced_table$chimp_bps_difs/reduced_table$human_length
  reduced_table$human_chimp_excess <- reduced_table$bps_difs - reduced_table$chimp_bps_difs
  reduced_table2 <- reduced_table %>% dplyr::filter(per_bp_difs < 0.75) %>% dplyr::filter(chimp_per_bp_difs < 0.75) 
  
  reduced_table3 <- reduced_table2 %>% 
    ungroup() %>% 
    dplyr::select(ID, human_length, bps_difs:chimp_bps_difs) %>% 
    group_by(ID) %>% 
    summarize(length = sum(human_length), human_fixed = sum(bps_difs), chimp_fixed = sum(chimp_bps_difs))
  
  write_delim(reduced_table3, , file = "~/Desktop/Capellini_Lab/Human_ATAC/fixed_bps_table.txt", delim = "\t")
  
  fixed_bp_table <- read.delim(file = "~/Desktop/Capellini_Lab/Human_ATAC/fixed_bps_table.txt", header = T, sep = "\t")
  
  
  #cleanup 
  system("rm ~/Desktop/TEMP_ATAC_DIR/temp_human_ape_bps.bed")
  
```

#repeat function and plotting for fixed bps

```{r}
#takes a df with two columns, first is the chr_start_end of the autopod only multint data, column 2 is the categorical data desired, column 3 is the lengths

divide_autopod_data_fixed_bp_plotting <- function(divisor_df, setID){
  #load bp data
    fixed_bp_table <- read.delim(file = "~/Desktop/Capellini_Lab/Human_ATAC/fixed_bps_table.txt", header = T, sep = "\t")
  
  #prepare dataframe to use for determining overlaps
  divisor_df2 <- divisor_df 
  colnames(divisor_df2) <- c("ID", "category", "size")
  divisor_df2 <- divisor_df2 %>% merge(y = fixed_bp_table, all.x =T, by = "ID")
  categories <- unique(divisor_df2$category)
  
    #create dataframe for results
  results_df <- data.frame(species = rep(c("human", "chimp"), length(categories)), 
                           set = setID, 
                           category = categories, 
                           value = NA, 
                           set_total = NA,
                           norm_value = NA)

  for (i in 1:length(categories)){
    results_df$set_total[which(results_df$category == categories[i])] <- divisor_df2 %>%  
      dplyr::filter(category == categories[i]) %>% 
      pull(size) %>% 
      sum(na.rm = T)
     results_df$value[which((results_df$category == categories[i]) & (results_df$species == "human"))] <- divisor_df2 %>%  
       dplyr::filter(category == categories[i]) %>% 
       pull(human_fixed) %>% 
      sum(na.rm = T)
     results_df$value[which((results_df$category == categories[i]) & (results_df$species == "chimp"))] <- divisor_df2 %>%  
       dplyr::filter(category == categories[i]) %>% 
       pull(chimp_fixed) %>% 
      sum(na.rm = T)
  }
  results_df$norm_value <- results_df$value/results_df$set_total
  return(results_df)
} #end function
```

#create a function to easily replicate figure for fig. 4, S8-9

```{r}
#takes a multiinter dataset
create_autopod_sharing_figure <- function(multiinter, col_end, filename, label){
  
  #add columns to multiinter needed for determining subsets
  multiinter <- add_multiinter_count_col(dataset = multiinter, grep_term = "Early", names = "Early", column_end = col_end)
  multiinter <- add_multiinter_count_col(dataset = multiinter, grep_term = "Late", names = "Late", column_end = col_end)
  
  multiinter  <- add_multiinter_count_col(dataset = multiinter, grep_term = "P", names = "Phalangeal", column_end = col_end)
  multiinter  <- add_multiinter_count_col(dataset = multiinter, grep_term = "M", names = "Metapodial", column_end = col_end)
  
  multiinter <- add_multiinter_count_col(dataset = multiinter, grep_term = "H", names = "Hand", column_end = col_end)
  multiinter <- add_multiinter_count_col(dataset = multiinter, grep_term = "F", names = "Foot", column_end = col_end)
  
  multiinter$stage_status <-"BOTH_STAGES"
  multiinter$stage_status[which(multiinter$Early > 0 & multiinter$Late == 0)] <- "EARLY"
  multiinter$stage_status[which(multiinter$Early == 0 & multiinter$Late > 0)] <- "LATE"
  
  multiinter$region_status <-"BOTH_REGIONS"
  multiinter$region_status[which(multiinter$Metapodial > 0 & multiinter$Phalangeal == 0)] <- "METAPODIAL"
  multiinter$region_status[which(multiinter$Metapodial == 0 & multiinter$Phalangeal > 0)] <- "PHALANGEAL"
  
  multiinter$limb_status <-"SHARED"
  multiinter$limb_status[which(multiinter$Foot > 0 & multiinter$Hand == 0)] <- "FOOT"
  multiinter$limb_status[which(multiinter$Foot == 0 & multiinter$Hand > 0)] <- "HAND"
  
  #create the constituent results dataframes 
hand_foot <- divide_autopod_only_data_plotting(
  multiinter %>% dplyr::select(ID, limb_status), 
  setID = "LIMB")
phalanges_metapodial <- divide_autopod_only_data_plotting(
  multiinter %>% dplyr::select(ID, region_status), 
  setID = "REGION")
early_late <- divide_autopod_only_data_plotting(
  multiinter %>% dplyr::select(ID, stage_status), 
  setID = "STAGE")

#merge the dataframes for one super plot
category_plotting_df <- rbind(hand_foot, phalanges_metapodial, early_late)
category_plotting_df$category <- gsub(pattern = "_", replacement = " ", category_plotting_df$category)
category_plotting_df$category <- factor(category_plotting_df$category, levels = c("FOOT", "SHARED", "HAND", "METAPODIAL", "BOTH REGIONS", "PHALANGEAL", "LATE", "BOTH STAGES", "EARLY"))
#clean up overlap names
category_plotting_df$overlap <- gsub(pattern = "HAR", replacement = "HARs", category_plotting_df$overlap)
category_plotting_df$overlap <- gsub(pattern = "HAQER", replacement = "HAQERs", category_plotting_df$overlap)
category_plotting_df$overlap <- gsub(pattern = "hCONDEL", replacement = "hCONDELs", category_plotting_df$overlap)
category_plotting_df$overlap <- gsub(pattern = "Inversion", replacement = "Inversions", category_plotting_df$overlap)
category_plotting_df$overlap <- gsub(pattern = "SDR", replacement = "SDRs", category_plotting_df$overlap)
category_plotting_df$overlap <- gsub(pattern = "human_SNPs", replacement = "Human SNPs", category_plotting_df$overlap)
category_plotting_df$overlap <- gsub(pattern = "chimp_SNPs", replacement = "Chimp SNPs", category_plotting_df$overlap)
category_plotting_df$overlap <- gsub(pattern = "gorilla_SNPs", replacement = "Gorilla SNPs", category_plotting_df$overlap)
category_plotting_df$overlap <- factor(category_plotting_df$overlap, levels = c("HARs", "HAQERs", "hCONDELs", "Inversions", "Inversion_Flank_500k", "Inversion_Flank_1M", "SDRs", "SDR_Flank_500k", "SDR_Flank_1M", "Human SNPs", "Chimp SNPs", "Gorilla SNPs"))

category_plotting_df$set <- factor(category_plotting_df$set, levels = c("STAGE", "REGION", "LIMB"))

#make barplot to show categories
categories_barplot <- ggplot(data = category_plotting_df %>% dplyr::filter(!overlap %in% c("Human SNPs", "Chimp SNPs", "Gorilla SNPs")) %>%  dplyr::select(set, category, set_total) %>% unique(), 
       aes(x = set, y = set_total, 
           fill = category)) + 
  geom_bar(stat = "identity") + 
  geom_text(aes(label = category), color = "white",
            position = position_stack(vjust = 0.5), size = 3) +
  coord_flip() +
  labs(y = label, color = "Set", x = NULL) + scale_fill_manual(
    values = c("#99CCFF", "#006699", "#003366", 
               "#99CC66", "#009933", "#006600", 
               "#FF99FF", "#CC0099", "#993399")) +
   theme_classic() +
  theme(legend.position = "none")

#plot categories scatterplot
category_plotting <- ggplot(data = category_plotting_df %>% dplyr::filter(overlap %in% c("HARs", "HAQERs", "hCONDELs", "Inversions", "SDRs", "Human SNPs", "Chimp SNPs", "Gorilla SNPs")), 
       aes(x = set, y = norm_value, 
           color = category)) + 
  geom_point(size = 2) +
  facet_wrap(overlap~ ., scales="free", ncol = 4) + 
  labs(y = "Overlaps per regulatory element", color = "Set", x = NULL) + scale_color_manual(
    values = c("#99CCFF", "#006699", "#003366", 
               "#99CC66", "#009933", "#006600", 
               "#FF99FF", "#CC0099", "#993399")) +
   theme_classic() +
  theme(legend.position = "none", axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        strip.background = element_blank(), 
        strip.text = element_text(face="bold", size = 6)) + 
  guides(color = guide_legend(ncol = 3, nrow = 3))

#stats plot

#create the constituent results dataframes 
hand_foot_stats <- divide_autopod_only_stat_plotting(
  multiinter %>% dplyr::select(ID, limb_status), 
  setID = "LIMB", exclude_SDR = T)

phalanges_metapodial_stats <- divide_autopod_only_stat_plotting(
  multiinter %>% dplyr::select(ID, region_status), 
  setID = "REGION", exclude_SDR = T)

early_late_stats <- divide_autopod_only_stat_plotting(
  multiinter %>% dplyr::select(ID, stage_status), 
  setID = "STAGE", exclude_SDR = T)


#merge the dataframes for one super plot
stat_plotting_df <- rbind(hand_foot_stats, phalanges_metapodial_stats, early_late_stats)

stat_plotting_df$comparison <- gsub(pattern = "_", replacement = " ", stat_plotting_df$comparison)
stat_plotting_df$comparison <- factor(stat_plotting_df$comparison, levels = c("FOOT vs SHARED", "HAND vs SHARED", "FOOT vs HAND", "PHALANGEAL vs BOTH REGIONS", "METAPODIAL vs BOTH REGIONS", "PHALANGEAL vs METAPODIAL", "EARLY vs BOTH STAGES", "LATE vs BOTH STAGES", "EARLY vs LATE"))

stat_plotting_df$overlap <- gsub(pattern = "HAR", replacement = "HARs", stat_plotting_df$overlap)
stat_plotting_df$overlap <- gsub(pattern = "HAQER", replacement = "HAQERs", stat_plotting_df$overlap)
stat_plotting_df$overlap <- gsub(pattern = "hCONDEL", replacement = "hCONDELs", stat_plotting_df$overlap)
stat_plotting_df$overlap <- gsub(pattern = "Inversion", replacement = "Inversions", stat_plotting_df$overlap)
stat_plotting_df$overlap <- gsub(pattern = "SDR", replacement = "SDRs", stat_plotting_df$overlap)
stat_plotting_df$overlap <- gsub(pattern = "human_SNPs", replacement = "Human SNPs", stat_plotting_df$overlap)
stat_plotting_df$overlap <- gsub(pattern = "chimp_SNPs", replacement = "Chimp SNPs", stat_plotting_df$overlap)
stat_plotting_df$overlap <- gsub(pattern = "gorilla_SNPs", replacement = "Gorilla SNPs", stat_plotting_df$overlap)
stat_plotting_df$overlap <- factor(stat_plotting_df$overlap, levels = c("HARs", "HAQERs", "hCONDELs", "Inversions", "Inversion_Flank_500k", "Inversion_Flank_1M", "SDRs", "SDR_Flank_500k", "SDR_Flank_1M", "Human SNPs", "Chimp SNPs", "Gorilla SNPs"))


stat_plotting_df$logP <- -log10(stat_plotting_df$pval)
stat_plotting_df$logP[which(stat_plotting_df$pval > 0.05)] <- NA

stat_plotting <- ggplot(data = stat_plotting_df %>% dplyr::filter(!overlap %in% c("SDR_Flank_500k", "SDR_Flank_1M", "Inversion_Flank_500k", "Inversion_Flank_1M"))) + 
  geom_point(size = 2, 
             data = stat_plotting_df %>% dplyr::filter(!overlap %in% c("SDR_Flank_500k", "SDR_Flank_1M", "Inversion_Flank_500k", "Inversion_Flank_1M")) %>% 
               dplyr::filter(!is.na(logP), test == "exact"), 
             aes(x = overlap, y = comparison, color = logP)) +
    scale_color_gradientn(colors = c( "lightgray", "gray9"), guide = guide_colorbar(order = 1)) +
  #facet_wrap(~set, nrow = 3) + 
  labs(color = "Fisher's Exact Test \n-log10(P-value)") +
  #start new scale
   new_scale_colour() +
  #apply the new gradient to color 2
  geom_point(size = 2, data = stat_plotting_df %>% dplyr::filter(!overlap %in% c("SDR_Flank_500k", "SDR_Flank_1M", "Inversion_Flank_500k", "Inversion_Flank_1M")) 
             %>% dplyr::filter(!is.na(logP), test == "wilcoxon"), 
             aes(x = overlap, y = comparison, 
           color = logP)) +
    scale_color_gradientn(colors = c("#fddbc7", "#b2182b"), guide = guide_colorbar(order = 2)) +
    labs(y = NULL, color = "Wilcoxon Rank Sum\n-log10(P-value)", x = NULL) +
   theme_classic() +
  theme(legend.position = "bottom", axis.title.x=element_blank(),
        strip.background = element_blank(), 
        strip.text = element_text(face="bold", size = 6)) + 
  scale_x_discrete(guide = guide_axis(angle = 45), drop=T) + scale_y_discrete(drop=T)

#create the constituent results dataframes 
hand_foot_fixed <- divide_autopod_data_fixed_bp_plotting(
  multiinter %>% 
    dplyr::mutate(size = end -start) %>% 
    dplyr::select(ID, limb_status, size), 
  setID = "LIMB")
phalanges_metapodial_fixed <- divide_autopod_data_fixed_bp_plotting(
  multiinter %>% 
    dplyr::mutate(size = end -start)
  %>% dplyr::select(ID, region_status, size), 
  setID = "REGION")
early_late_fixed <- divide_autopod_data_fixed_bp_plotting(
  multiinter %>% 
    dplyr::mutate(size = end -start)
  %>% dplyr::select(ID, stage_status, size), 
  setID = "STAGE")

#merge the dataframes for one super plot
fixed_plotting_df <- rbind(hand_foot_fixed, phalanges_metapodial_fixed, early_late_fixed)
fixed_plotting_df$category <- gsub(pattern = "_", replacement = " ", fixed_plotting_df$category)
fixed_plotting_df$category <- factor(fixed_plotting_df$category, levels = c("HAND", "SHARED", "FOOT", "PHALANGEAL", "BOTH REGIONS", "METAPODIAL", "EARLY", "BOTH STAGES", "LATE"))
fixed_plotting_df$species <- factor(fixed_plotting_df$species, levels = c("human", "chimp"))

fixed_plotting_df$set <- factor(fixed_plotting_df$set, levels = c("STAGE", "REGION", "LIMB"))


#plot results as paired bargraph
fixed_plotting <- ggplot(data = fixed_plotting_df, 
       aes(x = category, y = norm_value, 
           alpha = species, fill = category, color =category)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(y = "Fixed substitutions per bp", color = "Set", x = NULL) +
  scale_alpha_manual(values=c(1, 0)) + 
  scale_fill_manual(
    values = c("#003366", "#006699", "#99CCFF",
               "#006600", "#009933", "#99CC66",
                "#993399", "#CC0099", "#FF99FF")) + 
  scale_color_manual(
    values = c("#003366", "#006699", "#99CCFF",
               "#006600", "#009933", "#99CC66",
                "#993399", "#CC0099", "#FF99FF")) + 
  facet_wrap(~set, scales = "free_x") +
    geom_text(aes(label=value, group =species), vjust=-0.25, color="black",
            position = position_dodge(width = 0.9), size=2.5, alpha = 1) + 
   theme_classic() +
  theme(legend.position = "none", axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        strip.background = element_blank(), 
        strip.text = element_text(face="bold", size = 6), 
        strip.text.x = element_text(size=0)) + 
  guides(color = guide_legend(ncol = 3, nrow = 3))


categories_plot <- ggdraw() +
  draw_plot(categories_barplot, x = 0, y = 0.85, width = 1, height = .15) +
  draw_plot(category_plotting, x = 0, y = 0.55, width = 1, height = .3) +
  draw_plot(stat_plotting, x = 0, y = 0.25, width = 1, height = .3) +
  draw_plot(fixed_plotting, x = 0, y = 0, width = 1, height = .25) +
  draw_plot_label(label = c("a", "b", "c", "d"), 
                  size = 12,
                  x = c(0, 0, 0, 0 ), 
                  y = c(1, 0.85, 0.55, 0.25)) + 
  bgcolor("white")


ggsave(plot = categories_plot, filename = filename, device = "png", dpi = 300, height = 10, width = 8, units = "in")
  } #end function
```

#Create figure for all, autopod specific, mouse conserved

```{r}
#all
create_autopod_sharing_figure(multiinter = multiinter_both_df, 
                              col_end = 45, 
                              filename = "~/Dropbox/Autopod Paper/Autopod_Paper_Figures/all_categories_fig.png", 
                              label = "Number of regulatory elements (all)")

#autopod-specific
create_autopod_sharing_figure(multiinter = multiinter_autopod_only, 
                              col_end = 45, 
                              filename = "~/Dropbox/Autopod Paper/Autopod_Paper_Figures/figure_4_v3.png", 
                              label = "Number of regulatory elements (autopod-specific)")

#autopod-specific, mouse conserved
create_autopod_sharing_figure(multiinter = multiinter_autopod_only %>% dplyr::filter(ID %in% multiinter_mouse_IDs), 
                              col_end = 45, 
                              filename = "~/Dropbox/Autopod Paper/Autopod_Paper_Figures/mouse_categories_fig.png", 
                              label = "Number of regulatory elements (autopod-specific, mouse-conserved)")
```

#generate extended data fig. for flanking regions

```{r flanks}
#takes a multiinter dataset
create_autopod_flank_figure <- function(multiinter1, 
                                        multiinter2, 
                                        multiinter3,
                                        col_end, 
                                        filename, 
                                        filter_list, 
                                        mode = "plot"){
  
  #initialize dataframes for adding results
  category_plotting_df <- data.frame()
  stat_plotting_df <- data.frame()
  
  for(i in 1:3){
    if(i == 1){
        multiinter <- multiinter1
    }
    if(i == 2){
        multiinter <- multiinter2
    }
    if(i == 3){
        multiinter <- multiinter3
     }

  #add columns to multiinter needed for determining subsets
  multiinter <- add_multiinter_count_col(dataset = multiinter, grep_term = "Early", names = "Early", column_end = col_end)
  multiinter <- add_multiinter_count_col(dataset = multiinter, grep_term = "Late", names = "Late", column_end = col_end)
  
  multiinter  <- add_multiinter_count_col(dataset = multiinter, grep_term = "P", names = "Phalangeal", column_end = col_end)
  multiinter  <- add_multiinter_count_col(dataset = multiinter, grep_term = "M", names = "Metapodial", column_end = col_end)
  
  multiinter <- add_multiinter_count_col(dataset = multiinter, grep_term = "H", names = "Hand", column_end = col_end)
  multiinter <- add_multiinter_count_col(dataset = multiinter, grep_term = "F", names = "Foot", column_end = col_end)
  
  multiinter$stage_status <-"BOTH_STAGES"
  multiinter$stage_status[which(multiinter$Early > 0 & multiinter$Late == 0)] <- "EARLY"
  multiinter$stage_status[which(multiinter$Early ==0 & multiinter$Late > 0)] <- "LATE"
  
  multiinter$region_status <-"BOTH_REGIONS"
  multiinter$region_status[which(multiinter$Metapodial > 0 & multiinter$Phalangeal == 0)] <- "METAPODIAL"
  multiinter$region_status[which(multiinter$Metapodial == 0 & multiinter$Phalangeal > 0)] <- "PHALANGEAL"
  
  multiinter$limb_status <-"SHARED"
  multiinter$limb_status[which(multiinter$Foot > 0 & multiinter$Hand == 0)] <- "FOOT"
  multiinter$limb_status[which(multiinter$Foot == 0 & multiinter$Hand > 0)] <- "HAND"
  
  if(mode == "plot"){
        #create the constituent results dataframes 
      hand_foot <- divide_autopod_only_data_plotting(
        multiinter %>% dplyr::select(ID, limb_status), 
        setID = "LIMB")
      phalanges_metapodial <- divide_autopod_only_data_plotting(
        multiinter %>% dplyr::select(ID, region_status), 
        setID = "REGION")
      early_late <- divide_autopod_only_data_plotting(
        multiinter %>% dplyr::select(ID, stage_status), 
        setID = "STAGE")
      
      #merge the dataframes for one super plot
      category_plotting_df_temp <- rbind(hand_foot, phalanges_metapodial, early_late)
      category_plotting_df_temp$filtering <- filter_list[i]
      category_plotting_df <- rbind(category_plotting_df, category_plotting_df_temp)
    } #end plot mode option
    if(mode == "stats"){
      #stats plot

      #create the constituent results dataframes 
      hand_foot_stats <- divide_autopod_only_stat_plotting(
        multiinter %>% dplyr::select(ID, limb_status), 
        setID = "LIMB")

      phalanges_metapodial_stats <- divide_autopod_only_stat_plotting(
        multiinter %>% dplyr::select(ID, region_status), 
        setID = "REGION")
      
      early_late_stats <- divide_autopod_only_stat_plotting(
          multiinter %>% dplyr::select(ID, stage_status), 
        setID = "STAGE")


      #merge the dataframes for one super plot
      stat_plotting_df_temp <- rbind(hand_foot_stats, phalanges_metapodial_stats, early_late_stats)
      stat_plotting_df_temp$filtering <- filter_list[i]
      stat_plotting_df <- rbind(stat_plotting_df, stat_plotting_df_temp)
      
    } #end stats mode option
  } #end iterate over dataframes
  
  if(mode == "plot"){
      category_plotting_df$category <- gsub(pattern = "_", replacement = " ", category_plotting_df$category)
      category_plotting_df$category <- factor(category_plotting_df$category, levels = c("FOOT", "SHARED", "HAND", "METAPODIAL", "BOTH REGIONS", "PHALANGEAL", "LATE", "BOTH STAGES", "EARLY"))
      category_plotting_df$overlap <- gsub(pattern = "human_SNPs", replacement = "Human SNPs", category_plotting_df$overlap)
      category_plotting_df$overlap <- gsub(pattern = "chimp_SNPs", replacement = "Chimp SNPs", category_plotting_df$overlap)
      category_plotting_df$overlap <- gsub(pattern = "gorilla_SNPs", replacement = "Gorilla SNPs", category_plotting_df$overlap)
      category_plotting_df$overlap <- factor(category_plotting_df$overlap, levels = c("HAR", "HAQER", "hCONDEL", "Inversion", "Inversion_Flank_500k", "Inversion_Flank_1M", "SDR", "SDR_Flank_500k", "SDR_Flank_1M", "Human SNPs", "Chimp SNPs", "Gorilla SNPs"))
      
      category_plotting_df$set <- factor(category_plotting_df$set, levels = c("STAGE", "REGION", "LIMB"))
      return(category_plotting_df)
  }
  if(mode == "stats"){
      stat_plotting_df$comparison <- gsub(pattern = "_", replacement = " ", stat_plotting_df$comparison)
    stat_plotting_df$comparison <- factor(stat_plotting_df$comparison, levels = c("FOOT vs SHARED", "HAND vs SHARED", "FOOT vs HAND", "PHALANGEAL vs BOTH REGIONS", "METAPODIAL vs BOTH REGIONS", "PHALANGEAL vs METAPODIAL", "EARLY vs BOTH STAGES", "LATE vs BOTH STAGES", "EARLY vs LATE"))
    
    stat_plotting_df$overlap <- gsub(pattern = "human_SNPs", replacement = "Human SNPs", stat_plotting_df$overlap)
    stat_plotting_df$overlap <- gsub(pattern = "chimp_SNPs", replacement = "Chimp SNPs", stat_plotting_df$overlap)
    stat_plotting_df$overlap <- gsub(pattern = "gorilla_SNPs", replacement = "Gorilla SNPs", stat_plotting_df$overlap)
    stat_plotting_df$overlap <- factor(stat_plotting_df$overlap, levels = c("HAR", "HAQER", "hCONDEL", "Inversion", "Inversion_Flank_500k", "Inversion_Flank_1M", "SDR", "SDR_Flank_500k", "SDR_Flank_1M", "Human SNPs", "Chimp SNPs", "Gorilla SNPs"))
    
    stat_plotting_df$logP <- -log10(stat_plotting_df$pval)
    stat_plotting_df$logP[which(stat_plotting_df$pval > 0.05)] <- NA
    return(stat_plotting_df)
  }
}

#run flank function
flank_category_plotting_df <- create_autopod_flank_figure(multiinter1 = multiinter_both_df,
                            multiinter2 = multiinter_autopod_only,
                            multiinter3 = multiinter_autopod_only %>% dplyr::filter(ID %in% multiinter_mouse_IDs),
                              col_end = 45, 
                              filename = "~/Dropbox/Autopod Paper/Autopod_Paper_Figures/flank_figure.png",
                            filter_list = c("ALL", "AUTOPOD-SPECIFIC", "AUTOPOD-SPECIFIC, MOUSE-CONSERVED"))

flank_labels <- c(
  "SDR_Flank_500k" = "SDR Flank (500kb)", 
  "SDR_Flank_1M" = "SDR Flank (1Mb)", 
  "Inversion_Flank_500k" = "Inversion Flank (500kb)", 
  "Inversion_Flank_1M" = "Inversion Flank (1Mb)")

#plot categories scatterplot
flank_plotting_all <- ggplot(data = flank_category_plotting_df %>% dplyr::filter(overlap %in% c("SDR_Flank_500k", "SDR_Flank_1M", "Inversion_Flank_500k", "Inversion_Flank_1M")) %>% 
                               dplyr::filter(filtering == "ALL"), 
       aes(x = set, y = norm_value, 
           color = category)) + 
  geom_point(size = 2) +
  facet_wrap(overlap~ ., scales="free", labeller = as_labeller(flank_labels), ncol = 4) + 
  labs(y = "Overlaps per \nregulatory element", color = "Set", x = NULL, title = "All, brain-filtered") + scale_color_manual(
    values = c("#99CCFF", "#006699", "#003366", 
               "#99CC66", "#009933", "#006600", 
               "#FF99FF", "#CC0099", "#993399")) +
   theme_classic() +
  theme(legend.position = "none", axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        strip.background = element_blank(), 
        strip.text = element_text(face="bold", size = 6),
        plot.title = element_text(hjust = 0.5))


#plot categories scatterplot
flank_plotting_autopod <- ggplot(data = flank_category_plotting_df %>% dplyr::filter(overlap %in% c("SDR_Flank_500k", "SDR_Flank_1M", "Inversion_Flank_500k", "Inversion_Flank_1M")) %>% 
                               dplyr::filter(filtering == "AUTOPOD-SPECIFIC"), 
       aes(x = set, y = norm_value, 
           color = category)) + 
  geom_point(size = 2) +
  facet_wrap(overlap~ ., scales="free", labeller = as_labeller(flank_labels), ncol = 4) + 
  labs(y = "Overlaps per \nregulatory element", color = "Set", x = NULL, title = "Autopod-specific") + scale_color_manual(
    values = c("#99CCFF", "#006699", "#003366", 
               "#99CC66", "#009933", "#006600", 
               "#FF99FF", "#CC0099", "#993399")) +
   theme_classic() +
  theme(legend.position = "none", axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        strip.background = element_blank(), 
        strip.text = element_text(face="bold", size = 6),
        plot.title = element_text(hjust = 0.5))

#plot categories scatterplot
flank_plotting_mouse <- ggplot(data = flank_category_plotting_df %>% dplyr::filter(overlap %in% c("SDR_Flank_500k", "SDR_Flank_1M", "Inversion_Flank_500k", "Inversion_Flank_1M")) %>% 
                               dplyr::filter(filtering == "AUTOPOD-SPECIFIC, MOUSE-CONSERVED"), 
       aes(x = set, y = norm_value, 
           color = category)) + 
  geom_point(size = 2) +
  facet_wrap(overlap~ ., scales="free", labeller = as_labeller(flank_labels), ncol = 4) + 
  labs(y = "Overlaps per \nregulatory element", color = "Set", x = NULL, title = "Autopod-specific, human-mouse conserved") + scale_color_manual(
    values = c("#99CCFF", "#006699", "#003366", 
               "#99CC66", "#009933", "#006600", 
               "#FF99FF", "#CC0099", "#993399")) +
   theme_classic() +
  theme(legend.position = "bottom", axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        strip.background = element_blank(), 
        strip.text = element_text(face="bold", size = 6),
        plot.title = element_text(hjust = 0.5)) + 
  guides(color = guide_legend(ncol = 3, nrow = 3))

#get just the legend to ensure similar sizes
flank_legend <- get_legend(flank_plotting_mouse)

flank_plotting_mouse <- flank_plotting_mouse + theme(legend.position='none')

#now get the STATS
flank_stat_plotting_df <- create_autopod_flank_figure(multiinter1 = multiinter_both_df,
                            multiinter2 = multiinter_autopod_only,
                            multiinter3 = multiinter_autopod_only %>% dplyr::filter(ID %in% multiinter_mouse_IDs),
                              col_end = 45, 
                              filename = "~/Dropbox/Autopod Paper/Autopod_Paper_Figures/flank_figure.png",
                            filter_list = c("All", "Autopod-specific", "Autopod-specific, \nmouse-conserved"), 
                            mode = "stats")


flank_stat_plotting_df_clean <- flank_stat_plotting_df %>% dplyr::filter(overlap %in% c("SDR_Flank_500k", "SDR_Flank_1M", "Inversion_Flank_500k", "Inversion_Flank_1M"))
flank_stat_plotting_df_clean$overlap <- gsub(pattern = "_", replacement = " ", x = flank_stat_plotting_df_clean$overlap)
flank_stat_plotting_df_clean$overlap <- gsub(pattern = "500k", replacement = "(500kb)", x = flank_stat_plotting_df_clean$overlap)
flank_stat_plotting_df_clean$overlap <- gsub(pattern = "1M", replacement = "(1Mb)", x = flank_stat_plotting_df_clean$overlap)


flank_stat_plotting <- ggplot(flank_stat_plotting_df_clean %>% 
               dplyr::filter(!is.na(logP), test == "exact")) + 
  geom_point(size = 2, 
             aes(x = overlap, y = comparison, color = logP)) +
    scale_color_gradientn(colors = c( "lightgray", "gray9"), guide = guide_colorbar(order = 1)) +
  facet_wrap(~filtering, ncol = 3) + 
  labs(y = NULL, color = "Fisher's Exact Test \n-log10(P-value)") +
   theme_classic() +
  theme(legend.position = "bottom", axis.title.x=element_blank(),
        strip.background = element_blank(), 
        strip.text = element_text(face="bold", size = 6)) + 
  scale_x_discrete(guide = guide_axis(angle = 45), drop=T) + scale_y_discrete(drop=T)

#get just the legend to ensure similar sizes
stat_legend <- get_legend(flank_stat_plotting)

flank_stat_plotting <- flank_stat_plotting + theme(legend.position='none')

flank_plot <- ggdraw() +
  draw_plot(flank_plotting_all, x = 0, y = 0.8, width = 1, height = .2) +
  draw_plot(flank_plotting_autopod, x = 0, y = 0.6, width = 1, height = .2) +
  draw_plot(flank_plotting_mouse, x = 0, y = 0.4, width = 1, height = .2) +
  draw_plot(flank_legend, x = 0, y = 0.3, width = 0.5, height = .1) +
  draw_plot(stat_legend, x = 0.5, y = 0.3, width = 0.5, height = .1) +
  draw_plot(flank_stat_plotting, x = 0, y = 0, width = 1, height = .3) +
  draw_plot_label(label = c("a", "b", "c", "d"), 
                  size = 12,
                  x = c(0, 0, 0, 0 ), 
                  y = c(1, 0.8, 0.6, 0.3)) + 
  bgcolor("white")

ggsave(plot = flank_plot + panel_border(color = "black", size = 1), filename = "~/Dropbox/Autopod Paper/Autopod_Paper_Figures/S7_flank_figure.png", device = "png", dpi = 300, height = 10, width = 8, units = "in")
```

# create function to generate overlap figures (fig S5-6)
# first reformat data as needed

```{r}
multiinter_both_df2 <- multiinter_both_df %>% mutate(
  FM1 = pmax(FM1_Early, FM1_Late),
  FM2 = pmax(FM2_Early, FM2_Late),
  FM3 = pmax(FM3_Early, FM3_Late),
  FM4 = pmax(FM4_Early, FM4_Late),
  FM5 = pmax(FM5_Early, FM5_Late),
  FP1 = pmax(FP1_Early, FP1_Late),
  FP2 = pmax(FP2_Early, FP2_Late),
  FP3 = pmax(FP3_Early, FP3_Late),
  FP4 = pmax(FP4_Early, FP4_Late),
  FP5 = pmax(FP5_Early, FP5_Late),
  HM1 = pmax(HM1_Early, HM1_Late),
  HM2 = pmax(HM2_Early, HM2_Late),
  HM3 = pmax(HM3_Early, HM3_Late),
  HM4 = pmax(HM4_Early, HM4_Late),
  HM5 = pmax(HM5_Early, HM5_Late),
  HP1 = pmax(HP1_Early, HP1_Late),
  HP2 = pmax(HP2_Early, HP2_Late),
  HP3 = pmax(HP3_Early, HP3_Late),
  HP4 = pmax(HP4_Early, HP4_Late),
  HP5 = pmax(HP5_Early, HP5_Late),
) %>% 
  mutate(digit_count = rowSums(across(FM1:HP5)))

multiinter_autopod_only2 <- multiinter_autopod_only %>%  mutate(
  FM1 = pmax(FM1_Early, FM1_Late),
  FM2 = pmax(FM2_Early, FM2_Late),
  FM3 = pmax(FM3_Early, FM3_Late),
  FM4 = pmax(FM4_Early, FM4_Late),
  FM5 = pmax(FM5_Early, FM5_Late),
  FP1 = pmax(FP1_Early, FP1_Late),
  FP2 = pmax(FP2_Early, FP2_Late),
  FP3 = pmax(FP3_Early, FP3_Late),
  FP4 = pmax(FP4_Early, FP4_Late),
  FP5 = pmax(FP5_Early, FP5_Late),
  HM1 = pmax(HM1_Early, HM1_Late),
  HM2 = pmax(HM2_Early, HM2_Late),
  HM3 = pmax(HM3_Early, HM3_Late),
  HM4 = pmax(HM4_Early, HM4_Late),
  HM5 = pmax(HM5_Early, HM5_Late),
  HP1 = pmax(HP1_Early, HP1_Late),
  HP2 = pmax(HP2_Early, HP2_Late),
  HP3 = pmax(HP3_Early, HP3_Late),
  HP4 = pmax(HP4_Early, HP4_Late),
  HP5 = pmax(HP5_Early, HP5_Late),
)%>% 
  mutate(digit_count = rowSums(across(FM1:HP5)))

multiinter_mouse2 <- multiinter_mouse %>% mutate(
  FM1 = pmax(FM1E, FM1L),
  FM2 = pmax(FM2E, FM2L),
  FM3 = pmax(FM3E, FM3L),
  FM4 = pmax(FM4E, FM4L),
  FM5 = pmax(FM5E, FM5L),
  FP1 = pmax(FP1E, FP1L),
  FP2 = pmax(FP2E, FP2L),
  FP3 = pmax(FP3E, FP3L),
  FP4 = pmax(FP4E, FP4L),
  FP5 = pmax(FP5E, FP5L),
  HM1 = pmax(HM1E, HM1L),
  HM2 = pmax(HM2E, HM2L),
  HM3 = pmax(HM3E, HM3L),
  HM4 = pmax(HM4E, HM4L),
  HM5 = pmax(HM5E, HM5L),
  HP1 = pmax(HP1E, HP1L),
  HP2 = pmax(HP2E, HP2L),
  HP3 = pmax(HP3E, HP3L),
  HP4 = pmax(HP4E, HP4L),
  HP5 = pmax(HP5E, HP5L),
)

#now for early/late all
multiinter_late_df2 <- multiinter_both_df2[c(1:5, 26:46)] %>% 
  mutate(digit_count = rowSums(across(FM1_Late:HP5_Late))) %>% dplyr::filter(digit_count > 0)
names(multiinter_late_df2)[6:25] <- gsub(pattern = "_Late", replacement = "", names(multiinter_late_df2)[6:25])

multiinter_early_df2 <- multiinter_both_df2[c(1:25, 46)] %>% 
  mutate(digit_count = rowSums(across(FM1_Early:HP5_Early))) %>% dplyr::filter(digit_count > 0)

names(multiinter_early_df2)[6:25] <- gsub(pattern = "_Early", replacement = "", names(multiinter_early_df2)[6:25]) 

#autopod
multiinter_autopod_only2_late <- multiinter_autopod_only[c(1:5, 26:46)] %>% 
  mutate(digit_count = rowSums(across(FM1_Late:HP5_Late))) %>% dplyr::filter(digit_count > 0)
names(multiinter_autopod_only2_late)[6:25] <- gsub(pattern = "_Late", replacement = "", names(multiinter_autopod_only2_late)[6:25])

multiinter_autopod_only2_early <- multiinter_autopod_only[c(1:25, 46)] %>% 
  mutate(digit_count = rowSums(across(FM1_Early:HP5_Early))) %>% dplyr::filter(digit_count > 0)

names(multiinter_autopod_only2_early)[6:25] <- gsub(pattern = "_Early", replacement = "", names(multiinter_autopod_only2_early)[6:25]) 
```

# now create the function and run

```{r}
#plot takes three multiinter dataframes, one for the total set, one autopod specific, one autopod specific-mouse filtered

plot_evol_overlaps <- function(multiinter1, multiinter2, multiinter3, timepoint = "combined"){
  for (i in 1:3){
    #setting for total set
    if(i == 1){
      set = "all"
      multiinter <- multiinter1
    }
    #setting for autopod specific set
    if(i == 2){
      set = "autopod"
      multiinter <- multiinter2
    }
    #setting for autopod specific set
    if(i == 3){
      set = "specific"
      multiinter <- multiinter3
    }
    
  #plot total counts
  multiinter %>%
  #dplyr::filter(ID %in% XXX) %>% 
  dplyr::select(FM1:HP5) %>% 
  colSums() %>% 
  as.data.frame() %>% rename(value = names(.)[1]) %>% 
  tibble::rownames_to_column("tissue") %>%
  plot_autopod_expression(
    file_name = paste("all", set, timepoint, sep = "_"), 
    center = T
    )
  
  #plot HAR overlaps
  multiinter %>%
  dplyr::filter(ID %in% multiinter_HARs_IDs) %>% 
  dplyr::select(FM1:HP5) %>% 
  colSums() %>% 
  as.data.frame() %>% rename(value = names(.)[1]) %>% 
  tibble::rownames_to_column("tissue") %>%
  plot_autopod_expression(
    file_name = paste("hars", set, timepoint, sep = "_"),
    center = T
    )
  
  #plot HAQER overlaps
  multiinter %>%
  dplyr::filter(ID %in% multiinter_HAQERs_IDs) %>% 
  dplyr::select(FM1:HP5) %>% 
  colSums() %>% 
  as.data.frame() %>% rename(value = names(.)[1]) %>% 
  tibble::rownames_to_column("tissue") %>%
  plot_autopod_expression(
    file_name = paste("haqers", set, timepoint, sep = "_"),
    center = T
    )
  
  #plot hCONDEL overlaps
  multiinter %>%
  dplyr::filter(ID %in% multiinter_HARs_IDs) %>% 
  dplyr::select(FM1:HP5) %>% 
  colSums() %>% 
  as.data.frame() %>% rename(value = names(.)[1]) %>% 
  tibble::rownames_to_column("tissue") %>%
  plot_autopod_expression(
    file_name = paste("hcondels", set, timepoint, sep = "_"),
    center = T
    )
  
  #plot inversion overlaps
  multiinter %>%
  dplyr::filter(ID %in% multiinter_inversions_IDs) %>% 
  dplyr::select(FM1:HP5) %>% 
  colSums() %>% 
  as.data.frame() %>% rename(value = names(.)[1]) %>% 
  tibble::rownames_to_column("tissue") %>%
  plot_autopod_expression(
    file_name = paste("inversions", set, timepoint, sep = "_"),
    center = T
    )
  
  #plot SDRs overlaps
  multiinter %>%
  dplyr::filter(ID %in% multiinter_SDRs_IDs) %>% 
  dplyr::select(FM1:HP5) %>% 
  colSums() %>% 
  as.data.frame() %>% rename(value = names(.)[1]) %>% 
  tibble::rownames_to_column("tissue") %>%
  plot_autopod_expression(
    file_name = paste("sdrs", set, timepoint, sep = "_"),
    center = T
    )
  
  } #end loop over multiinters
}

plot_evol_overlaps(multiinter1 = multiinter_both_df2, multiinter2 = multiinter_autopod_only2, multiinter3 = multiinter_autopod_only2 %>% dplyr::filter(digit_count == 1))

#early
plot_evol_overlaps(multiinter1 = multiinter_early_df2, multiinter2 = multiinter_autopod_only2_early, multiinter3 = multiinter_autopod_only2_early %>% dplyr::filter(digit_count == 1), "early")

#late
plot_evol_overlaps(multiinter1 = multiinter_late_df2, multiinter2 = multiinter_autopod_only2_late, multiinter3 = multiinter_autopod_only2_late %>% dplyr::filter(digit_count == 1), timepoint = "late")
```
