---
title: "Human Autopod RNA Analysis"
author: "Alexander Okamoto"
date: "2025-12-10"
output: html_document
---

##Set working directory
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, root.dir = '~/Desktop/Capellini_Lab/Human Autopod RNA/Human RSEM Output CORRECTED/')
```

##load required packages

```{r, error=FALSE, warning = FALSE}
library(tidyverse) #for basic data manipulation
library(DESeq2) #for RNA differential expression
library(limma) #for batch effect correction
library(data.table) #for data manipulation
library(WGCNA) #for cluster analysis
library(grid) # for plotting
library(data.table) #for data manipulation
library(ggpubr) #for making combined figures
library(cowplot) #for making combined figures
library(magick) #for loading pictures into ggplot

library(stringr) # for manipulating strings
library(clusterProfiler) #for gene enrichment
library(org.Hs.eg.db) #for gene enrichment
library(org.Mm.eg.db) #for gene enrichment


#load custom functions for analysis
source("~/Desktop/Capellini_Lab/Weekly_Coding/RNA_Analysis_Functions.R") 
```

## Load in the data

```{r}

#Determine the file names
sampleNames <- read.table(file = "~/Desktop/Capellini_Lab/Human Autopod RNA/Human RSEM Output CORRECTED/human_autopod_RNA_samples.txt", header = F)
#sort sample names alphabetically 
sampleNames <- sort(sampleNames$V1)

#create dataframe including metadata information
sampleMetaData <- data.frame(sample = sampleNames, limbtype=c(rep(c("Hind_limb"), 10), rep(c("Forelimb"), 10)), digit=c("I", "II", "III", "IV", "V"), tissue = c(rep("Metapodial", 5), rep("Phalange", 5), rep("Metapodial", 5), rep("Phalange", 5)), tissue2 = c(rep("Metatarsal", 5), rep("Hind_phalange", 5), rep("Metacarpal", 5), rep("Fore_phalange", 5)), replicate = c(rep('H28662', 20), rep('H28670', 20), rep('H28693A', 20), rep('H28693B', 20), rep('H28743', 20), rep('H28769', 20), rep('H28841', 20), rep('H28852', 20), rep('H28853', 20), rep('H28867', 20), rep('H28906', 20), rep('H29057', 20)), sampleType = c("FM1", "FM2", "FM3", "FM4", "FM5", "FP1", "FP2", "FP3", "FP4", "FP5", "HM1", "HM2", "HM3", "HM4", "HM5", "HP1", "HP2", "HP3", "HP4", "HP5"), sex = c(rep("Male", 100), rep("Female", 20), rep("Male", 40), rep("Female", 20), rep("Male", 60)), timepoint = c(rep('Late', 100), rep('Early', 40), rep('Late', 20), rep('Early', 80)), stage = c(rep('72', 20), rep('67', 20), rep('74', 20), rep('72', 20), rep('67', 20), rep('57', 20), rep('53', 20), rep('72', 20), rep('59', 20), rep('54', 20), rep('57', 20), rep('54', 20)))

#remove problematic samples
sampleMetaData <- sampleMetaData %>% dplyr::filter(!sample %in% c("28867_HM1", "29057_FM3"))

rownames(sampleMetaData) <- sampleMetaData$sample

#change variables to factors
sampleMetaData$limbtype <- as.factor(sampleMetaData$limbtype)
sampleMetaData$replicate <- as.factor(sampleMetaData$replicate)
sampleMetaData$sex <- as.factor(sampleMetaData$sex)
sampleMetaData$timepoint <- as.factor(sampleMetaData$timepoint)
sampleMetaData$stage <- as.factor(sampleMetaData$stage)
sampleMetaData$sample_tp <- paste(sampleMetaData$sampleType, "_", sampleMetaData$timepoint, sep = "")

#get load read count information to check for technical artifacts in the data
human_rna_read_counts <- read.delim(file = "~/Desktop/Capellini_Lab/Human Autopod RNA/human_rna_read_counts.txt", header = T)
#rename columns for clarity/merging
colnames(human_rna_read_counts) <- c("sample", "reads")

#add read count information to metadata dataframe
sampleMetaData <- sampleMetaData %>% dplyr::left_join(human_rna_read_counts, by = 'sample')

#load the counts data
ct <- read_in_counts(samplenames = sampleMetaData$sample, filepath = "~/Desktop/Capellini_Lab/Human Autopod RNA/Human RSEM Output CORRECTED/")

#remove version number from gene_id so it will be compatible with biomart
ct$gene_id = sapply(strsplit(ct$gene_id, ".", fixed=T), function(x) x[1])

#will sum the counts for matches to the same gene
ct <- aggregate(. ~ gene_id, data= ct, FUN=sum)

#change row names to unique values
rownames(ct) <- ct$gene_id
names(ct)[1] <- "ensembl_gene_id"

#remove samples with low read counts
ct$`28867_HM1` <- NULL
ct$`29057_FM3` <- NULL
```

##RSEM output human ensembl ids with the version number (with .X) and biomart/DEGseq would prefer they be normal integers. The problem is some of these values are non-unique so need to figure out what to do with these. There are only 33 duplicated genes so these will be aggregrated. 

##load in ENSEMBL human gene data

```{r pressure, echo=FALSE}
require(biomaRt)
hs_ensembl <- useMart('ensembl', dataset = 'hsapiens_gene_ensembl', host = "https://asia.ensembl.org")

#make annotation data frame
annot <- getBM(
  attributes = c(
    'hgnc_symbol',
    'ensembl_gene_id',
    'gene_biotype'),
  filters = 'ensembl_gene_id',
  values = ct$ensembl_gene_id,
  mart = hs_ensembl)

#merge annotations to counts data
annot_df <- merge(
  x = as.data.frame(ct),
  y =  annot,
  by = 'ensembl_gene_id',
  all.x = T)

#cleanup
rm(hs_ensembl)
```

## Determine pair-wise comparisons

```{r}
## now determine the specific pairwise comparisons to perform (i.e. the most biologically meaningful DESeq runs)
#I am using a schema to minimize confounders by only changing a single variable (either tissue location, either anteriorly/posteriorly or proximally/distally, across limb type, or across timepoint).

#list sample types
sampleTypes <- c(paste(unique(sampleMetaData$sampleType), "E", sep = ""), paste(unique(sampleMetaData$sampleType), "L", sep = ""))

#create dataframe to store results
sample_unique_info <- data.frame(sample = sampleTypes, comparisons1 = NA, comparisons2= NA)

#create sample and control sets
for (samp in sampleTypes){
  comp_digits <- list()
  alt_region <- "Unknown"
  digit_comp <- "Unknown"
  neighbors <- "Unknown"
  samp_info <-unlist(strsplit(samp, ""))
  #check if digit 1
  if (samp_info[3] == "1"){
    comp_digits <- 2
  }
  #check if digit 3
  if (samp_info[3] > 1 && samp_info[3] < 5){
    comp_digits <- c(as.numeric(samp_info[3]) - 1,as.numeric(samp_info[3]) + 1)
  }
  #check if digit 1
  if (samp_info[3] == "5"){
    comp_digits <- 4
  }
  #Get other region
  #check if metapodial
  if (samp_info[2] == "M"){
    alt_region <- "P"
  }
  #check if phalanges
  if (samp_info[2] == "P"){
    alt_region <- "M"
  }
  #check if hand (forelimb)
  if (samp_info[1] == "H"){
    alt_limb <- "F"
  }
  #check if food (hind limb)
  if (samp_info[1] == "F"){
    alt_limb <- "H"
  }
  #check if early stage
  if (samp_info[4] == "E"){
    alt_stage <- "L"
  }
  #check if late stage
  if (samp_info[4] == "L"){
    alt_stage <- "E"
  }
    digit_comp <- paste(samp_info[1], alt_region, samp_info[3], samp_info[4], sep = "")
    neighbors <- list()
    for (n in 1:length(comp_digits)){
      neighbors <- append(neighbors, paste(samp_info[1], samp_info[2], comp_digits[n], samp_info[4], sep = ""))
    }
  neighbors2 <- paste(neighbors, collapse ="|")
  limb_comp <- paste(alt_limb, samp_info[2], samp_info[3], samp_info[4], sep = "")
  stage_comp <- paste(samp_info[1], samp_info[2], samp_info[3], alt_stage, sep = "")
  sample_unique_info$comparisons1[which(sample_unique_info$sample == samp) ] <- list(as.list(c(neighbors, digit_comp, limb_comp, stage_comp)))
  sample_unique_info$comparisons2[which(sample_unique_info$sample == samp) ] <- paste(samp, neighbors2, digit_comp, limb_comp, stage_comp, sep = "|")
}
```

## Create a function to perform DESeq and return number of genes

```{r}
#initialize function
  #design_in is the DESeq model input

runDESEQ2_df_hs <- function(design_in, count_data, col_data, padj.cutoff = 0.05){

#glm design
# Again, we can include more things beside tissue, like we could look at the effect of tissue *and* age, or we could control for sex here. 
  
#build DESeqDataSet
rsem.in <- DESeqDataSetFromMatrix(countData = count_data, colData = col_data, design = design_in, tidy = T)
#pre-filter rows with 0 or 1 read
rsem.in <- rsem.in[ rowSums(counts(rsem.in)) >= 1, ]
rsem_cols <-as.data.frame(row.names(rsem.in))
colnames(rsem_cols) <- 'ensembl_gene_id'
rsem_annot <- merge(x= rsem_cols, y =  annot,
  by= 'ensembl_gene_id', no.dups = TRUE, all.x=TRUE) %>% setorder(cols = "ensembl_gene_id")  
#remove duplicate rows
rsem_annot <- rsem_annot[!duplicated(rsem_annot$ensembl_gene_id),]
mcols(rsem.in) <- cbind(mcols(rsem.in), rsem_annot$hgnc_symbol)
rsem.de <- DESeq(rsem.in)

#rsem.in variable is building the dataset 
#now model is in rsem.de

#summarized experiment input
#can set alpha level at whatever
rsem.de.res <- results(rsem.de, alpha = 0.05)
#
rsem.de.res$mgi_symbol <- mcols(rsem.in)['rsem_annot$hgnc_symbol']

#summary shows summary of results
summary(rsem.de.res)

#This is a way to export a table of the differentially expressed genes, ordered by p-value
rsem.de.res.df <- rsem.de.res[order(rsem.de.res$padj),] %>% as.data.frame() %>% dplyr::filter(padj < padj.cutoff)

return(rsem.de.res.df)
} #end function

#add key to convert between my nomenclature and something less easy to misinterpret
autopod_key_notp <- data.frame(autopod_abrevs = c("FM1", "FM2", "FM3", "FM4", "FM5", 
                    "FP1", "FP2", "FP3", "FP4", "FP5", 
                    "HM1", "HM2", "HM3", "HM4", "HM5", 
                    "HP1", "HP2", "HP3", "HP4", "HP5"), autopod_tissues = c("METATARSAL1", "METATARSAL2", "METATARSAL3", "METATARSAL4", "METATARSAL5", 
                                   "FOOTPHALANGES1", "FOOTPHALANGES2", "FOOTPHALANGES3", "FOOTPHALANGES4", "FOOTPHALANGES5", 
                                   "METACARPAL1", "METACARPAL2", "METACARPAL3", "METACARPAL4", "METACARPAL5", 
                                   "HANDPHALANGES1", "HANDPHALANGES2", "HANDPHALANGES3", "HANDPHALANGES4", "HANDPHALANGES5"))
autopod_key_early <- autopod_key_notp 
autopod_key_late <- autopod_key_notp 
autopod_key_early$autopod_abrevs <- paste(autopod_key_early$autopod_abrevs, "E", sep = "")
autopod_key_late$autopod_abrevs <- paste(autopod_key_late$autopod_abrevs, "L", sep = "")
autopod_key_early$autopod_tissues <- paste(autopod_key_early$autopod_tissues, "E", sep = "")
autopod_key_late$autopod_tissues <- paste(autopod_key_late$autopod_tissues, "L", sep = "")
  
autopod_key <- rbind(autopod_key_early, autopod_key_late)
  

#modify ct column names to be compatible
names(ct)[which(names(ct) %in% sampleMetaData$sample[which(sampleMetaData$timepoint == "Late")])] <- paste(names(ct)[which(names(ct) %in% sampleMetaData$sample[which(sampleMetaData$timepoint == "Late")])], "L", sep = "")
names(ct)[which(names(ct) %in% sampleMetaData$sample[which(sampleMetaData$timepoint == "Early")])] <- paste(names(ct)[which(names(ct) %in% sampleMetaData$sample[which(sampleMetaData$timepoint == "Early")])], "E", sep = "")

#add column to sample metadata
sampleMetaData$sampleTypeTP <- NA
sampleMetaData$sampleTypeTP[which(sampleMetaData$timepoint == "Late")] <- paste(sampleMetaData$sampleType[which(sampleMetaData$timepoint == "Late")], "L", sep = "")
sampleMetaData$sampleTypeTP[which(sampleMetaData$timepoint == "Early")] <- paste(sampleMetaData$sampleType[which(sampleMetaData$timepoint == "Early")], "E", sep = "")
sampleMetaData$sampleTypeTP<- factor(sampleMetaData$sampleTypeTP)
rownames(sampleMetaData)[which(sampleMetaData$timepoint == "Late")] <- paste(sampleMetaData$sample[which(sampleMetaData$timepoint == "Late")], "L", sep = "")
rownames(sampleMetaData)[which(sampleMetaData$timepoint == "Early")] <- paste(sampleMetaData$sample[which(sampleMetaData$timepoint == "Early")], "E", sep = "")

#add rownames to sampleMetaData
row.names(sampleMetaData) <- names(ct[-1])
# #create dataframe to store results
 human_DEGs <- data.frame()
#
 #iterate over each sample
 for (i in 1:length(sampleTypes)){
   #find number of
   comp_samples <- unlist(sample_unique_info$comparisons1[i])
   for(c in 1:length(comp_samples)){
     #filter data and metadata
     colsToKeep <- c(1, pattern = grep(paste(sample_unique_info$sample[i], comp_samples[c], sep = "|"), names(ct)))
     good_data <- ct[,colsToKeep]
     filteredMetaData <- sampleMetaData %>% dplyr::filter(sampleTypeTP %in% c(sample_unique_info$sample[i], comp_samples[c]))
filteredMetaData$sampleTypeTP <- factor(filteredMetaData$sampleTypeTP, levels = c(sample_unique_info$sample[i], comp_samples[c]))
     #check if same all same timepoint
     if(length(names(good_data[-1])) ==  length(grep(x = names(good_data[-1]), pattern = "L"))){
     test_output <- runDESEQ2_df_hs(design_in = formula(~replicate + sampleTypeTP), count_data = good_data, col_data = filteredMetaData) %>% setDT(keep.rownames = T)
     }else{
       test_output <- runDESEQ2_df_hs(design_in = formula(~sampleTypeTP), count_data = good_data, col_data = filteredMetaData) %>% setDT(keep.rownames = T)
     }
     #fix column names
     names(test_output)[1] <- "ensembl_gene_id"
     names(test_output)[8] <- "hgnc_symbol"
     test_output$tissue1 <- autopod_key$autopod_tissues[which(autopod_key$autopod_abrevs == sample_unique_info$sample[i])]
     test_output$tissue2 <- autopod_key$autopod_tissues[which(autopod_key$autopod_abrevs == comp_samples[c])]
     if(nrow(test_output) > 0){
     human_DEGs <- rbind(human_DEGs, test_output)
     }
     #note if no DEGs identified
     if(nrow(test_output) == 0){
     test_output <- data.frame(ensembl_gene_id = NA,  baseMean = NA, log2FoldChange = NA, lfcSE = NA, stat = NA, pvalue= NA, padj = NA, hgnc_symbol= "NO_DEGs_FOUND", tissue1=autopod_key$autopod_tissues[which(autopod_key$autopod_abrevs == sample_unique_info$sample[i])], tissue2= autopod_key$autopod_tissues[which(autopod_key$autopod_abrevs == comp_samples[c])])
     human_DEGs <- rbind(human_DEGs, test_output)
     }
   }
 }

 write_delim(x = as.data.frame(human_DEGs), file= "/Users/alexanderokamoto/Desktop/Capellini_Lab/Human Autopod RNA/All_Human_Bio_Pairwise_Autopod_DEGs.tsv", delim = "\t")
```

## Gradient genes

```{r}
#subset the count data
colnames_clean <- gsub(pattern = "E", replacement = "", colnames(ct))
colnames_clean <- gsub(pattern = "L", replacement = "", colnames_clean)

#early
ct_early <- ct[, c(1, which(colnames_clean %in% sampleMetaData$sample[which(sampleMetaData$timepoint == "Early")]))] %>% as.data.frame()
#pre-filter rows with 0 or 1 read
ct_early <- ct_early[rowSums(ct_early[,-1]) > 1,]

#late
ct_late <- ct[,c(1, which(colnames_clean %in% sampleMetaData$sample[which(sampleMetaData$timepoint == "Late")]))]
#pre-filter rows with 0 or 1 read
ct_late <- ct_late[rowSums(ct_late[-1]) > 1,]

#now identify gradient genes
dds <- DESeqDataSetFromMatrix(countData = ct_early, colData = sampleMetaData %>% dplyr::filter(timepoint == "Early"), design = ~replicate + digit, tidy = T)
dds <- DESeq(dds, test="LRT", reduced=~replicate)
dds_res <- results(dds, alpha = 0.05)
dds_res_df_early <- merge(x= as.data.frame(dds_res), by.x ='row.names',  y =  annot,
  by.y = 'ensembl_gene_id', no.dups = TRUE) %>% dplyr::filter(padj < 0.05)

#reform to add to human DEG object
colnames(dds_res_df_early)[1] <- "ensembl_gene_id"
dds_res_df_early$tissue1 <- "EARLY_GRADIENT"
dds_res_df_early$tissue2 <- "EARLY_GRADIENT"
dds_res_df_early$gene_biotype <- NULL

human_DEGs2 <- rbind(human_DEGs, dds_res_df_early)

#now identify gradient genes
dds <- DESeqDataSetFromMatrix(countData = ct_late, colData = sampleMetaData %>% dplyr::filter(timepoint == "Late"), design = ~replicate + digit, tidy = T)
dds <- DESeq(dds, test="LRT", reduced=~replicate)
dds_res <- results(dds, alpha = 0.05)
dds_res_df_late <- merge(x= as.data.frame(dds_res), by.x ='row.names',  y =  annot,
  by.y = 'ensembl_gene_id', no.dups = TRUE) %>% dplyr::filter(padj < 0.05)

#reform to add to human DEG object
colnames(dds_res_df_late)[1] <- "ensembl_gene_id"
dds_res_df_late$tissue1 <- "LATE_GRADIENT"
dds_res_df_late$tissue2 <- "LATE_GRADIENT"
dds_res_df_late$gene_biotype <- NULL

#get final set of human DEGs
human_DEGs2 <- rbind(human_DEGs2, dds_res_df_late) %>% dplyr::filter(padj < 0.05, abs(log2FoldChange) > 0.58)

  write_delim(x = as.data.frame(human_DEGs2), file= "/Users/alexanderokamoto/Desktop/Capellini_Lab/Human Autopod RNA/All_Human_Bio_Pairwise_Autopod_DEGs2.tsv", delim = "\t")
```

## Get timepoint lists of DEGs

```{r}
human_DEGs2 <- read_table(file= "/Users/alexanderokamoto/Desktop/Capellini_Lab/Human Autopod RNA/All_Human_Bio_Pairwise_Autopod_DEGs.tsv")

#get timepoint information for each DEG comparison
human_DEGs2$tissue1_tp <- str_sub(human_DEGs2$tissue1, start= -1)
human_DEGs2$tissue2_tp <- str_sub(human_DEGs2$tissue2, start= -1)
#fix gradient gene calls
human_DEGs2$tissue1_tp[which(human_DEGs2$tissue1 == "EARLY_GRADIENT")] <- "E"
human_DEGs2$tissue2_tp[which(human_DEGs2$tissue2 == "EARLY_GRADIENT")] <- "E"
human_DEGs2$tissue1_tp[which(human_DEGs2$tissue1 == "LATE_GRADIENT")] <- "L"
human_DEGs2$tissue2_tp[which(human_DEGs2$tissue2 == "LATE_GRADIENT")] <- "L"

#subset the DEGs
human_DEGs_early <- human_DEGs2 %>% dplyr::filter(tissue1_tp == "E" & tissue2_tp == "E")
human_DEGs_late <- human_DEGs2 %>% dplyr::filter(tissue1_tp == "L" & tissue2_tp == "L")

#get counts
length(unique(human_DEGs_early$ensembl_gene_id))
length(unique(human_DEGs_late$ensembl_gene_id))
length(intersect(human_DEGs_early$ensembl_gene_id, human_DEGs_late$ensembl_gene_id))


length(unique(human_DEGs2$ensembl_gene_id))
human_DEGs2 %>% dplyr::filter(tissue1 %in% c("EARLY_GRADIENT", "LATE_GRADIENT")) %>% pull(ensembl_gene_id) %>% unique %>% length()
human_DEGs2 %>% dplyr::filter(! tissue1 %in% c("EARLY_GRADIENT", "LATE_GRADIENT")) %>% pull(ensembl_gene_id) %>% unique %>% length()

```

## Module Analysis Using WGNCA
I used this tutoriaL: https://bioinformaticsworkbook.org/tutorials/wgcna.html#gsc.tab=0

#rerun a DEGseq to get vst transformed data
#forelimb vs hind limb
#early

```{r}
#glm design
# Again, we can include more things beside tissue, like we could look at the effect of tissue *and* age, or we could control for sex here. 
design <- formula(~replicate + limbtype)
#build DESeqDataSet
rsem.in <- DESeqDataSetFromMatrix(countData = ct_early, colData = sampleMetaData %>% dplyr::filter(timepoint == "Early"), design = design, tidy = T)

rsem_cols <-as.data.frame(row.names(rsem.in))
rsem_cols$ensembl <- sapply(strsplit(rsem_cols$`row.names(rsem.in)`, ".", fixed=T), function(x) x[1])

colnames(rsem_cols) <- c('ensembl_gene_id', 'ensembl')
rsem_annot <- merge(x= rsem_cols, y =  annot, by.x ='ensembl',
  by.y = 'ensembl_gene_id', no.dups = FALSE, all.x = TRUE, all.y=FALSE)

#combine gene symbols for duplicated rows
for (dup in rsem_annot$ensembl_gene_id[which(duplicated(rsem_annot$ensembl_gene_id))]){
  dup_ids <- which(rsem_annot$ensembl_gene_id == dup)
  dup_cat_id <- paste(rsem_annot$ensembl_gene_id[1], ", ", rsem_annot$ensembl_gene_id[2], sep ="")
  rsem_annot$hgnc_symbol[dup_ids[2]]<- dup_cat_id
}
rsem_annot_2 <- rsem_annot[-which(duplicated(rsem_annot$ensembl_gene_id)), ] %>% setorder(cols = "ensembl_gene_id")   
mcols(rsem.in) <- cbind(mcols(rsem.in), rsem_annot_2$hgnc_symbol)

#, test = "LRT", reduced = reducedest--pairwise does not need it. Or post-hoc tests can compare between pairs. 
rsem.de <- DESeq(rsem.in)
mcols(rsem.de) <- cbind(mcols(rsem.de), rsem_annot_2)

# Retrieve the normalized data from the `DESeqDataSet`

#plots related to dispersion, built in function to DESEQ
rsem.vst_early <- vst(rsem.de)
mat <- assay(rsem.vst_early)
mm <- model.matrix(~replicate, colData(rsem.vst_early))
mat <- limma::removeBatchEffect(mat, batch=colData(rsem.vst_early)$replicate)
assay(rsem.vst_early) <- mat

normalized_counts_early <- assay(rsem.vst_early)

#filter by DEGs
human_DEGs_early_list <- unique(human_DEGs_early$ensembl_gene_id)
DEG_normalized_counts_early <- normalized_counts_early[which(row.names(normalized_counts_early) %in% human_DEGs_early_list),]

#library(WGCNA)
input_mat = t(DEG_normalized_counts_early)

# Choose a set of soft-thresholding powers
powers = c(c(1:10), seq(from = 12, to = 30, by = 2))

# Call the network topology analysis function
sft = pickSoftThreshold(
  input_mat,             # <= Input data
  #blockSize = 30,
  powerVector = powers,
  verbose = 5
  )

par(mfrow = c(1,2));
cex1 = 0.9;

plot(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     xlab = "Soft Threshold (power)",
     ylab = "Scale Free Topology Model Fit, signed R^2",
     main = paste("Scale independence")
)
text(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     labels = powers, cex = cex1, col = "red"
)
abline(h = 0.90, col = "red")
plot(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     xlab = "Soft Threshold (power)",
     ylab = "Mean Connectivity",
     type = "n",
     main = paste("Mean connectivity")
)
text(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     labels = powers,
     cex = cex1, col = "red")

picked_power = min(which(sft$fitIndices$SFT.R.sq > 0.9))
temp_cor <- cor       
cor <- WGCNA::cor         # Force it to use WGCNA cor function (fix a namespace conflict issue)
netwk <- blockwiseModules(input_mat,                # <= input here

                          # == Adjacency Function ==
                          power = picked_power,                # <= power here
                          networkType = "signed",

                          # == Tree and Block Options ==
                          deepSplit = 2,
                          pamRespectsDendro = F,
                          # detectCutHeight = 0.75,
                          minModuleSize = 30,
                          maxBlockSize = 4000,

                          # == Module Adjustments ==
                          reassignThreshold = 0,
                          mergeCutHeight = 0.25,

                          # == TOM == Archive the run results in TOM file (saves time)
                          saveTOMs = T,
                          saveTOMFileBase = "ER",

                          # == Output Options
                          numericLabels = T,
                          verbose = 3)

module_df_early <- data.frame(
  gene_id = names(netwk$colors),
  cluster = netwk$colors,
  color = labels2colors(netwk$colors)
)


# Get Module Eigengenes per cluster
MEs <- moduleEigengenes(input_mat, netwk$colors)$eigengenes
MEs0 <- MEs

# Reorder modules so similar modules are next to each other
#MEs0 <- orderMEs(MEs0)
module_order = names(MEs0) %>% gsub("ME","", .)

# Add treatment names
MEs0$treatment <- sampleMetaData %>% dplyr::filter(timepoint == "Early") %>% pull(sampleType)

# tidy & plot data
mME = MEs0 %>%
  pivot_longer(-treatment) %>%
  mutate(
    color = gsub("ME", "", name),
    color = factor(color, levels = module_order) 
  ) %>% 
  group_by(treatment, color) %>% 
  summarize(avg_val = mean(value))

#create key to convert between cluster color and number
#Note: The grey color/module 0 is reserved for unassigned genes
color_cluster_key <- module_df_early %>% 
  group_by(cluster) %>% 
  summarize(gene_n = n()) 

write_delim(module_df_early %>% left_join(color_cluster_key) %>% 
  left_join(annot, by = join_by(gene_id == ensembl_gene_id)) %>% dplyr::select(gene_id, hgnc_symbol, cluster) %>% dplyr::rename(ensembl_gene_id= gene_id) %>% arrange(cluster),
            file = "~/Desktop/Capellini_Lab/Human Autopod RNA/human_early_gene_modules.tsv",
            delim = "\t")

for(i in min(color_cluster_key$cluster):max(color_cluster_key$cluster)){

  mME_temp <- mME %>% mutate(cluster = as.numeric(unfactor(color))) %>% 
    left_join(color_cluster_key) %>% 
    dplyr::filter(cluster == i) %>% 
                          mutate(value = avg_val) %>% 
                          mutate(tissue = treatment)
  
plot_autopod_expression(mME_temp %>% 
                          dplyr::select(tissue, value), 
                        file_name = paste("human_early_RNA_cluster_", i, sep = ""), opt_text = paste(mME_temp$gene_n[i+1], " genes", sep =""),
                       center =T 
                        )
  
}

plt0 <- image_read("~/Desktop/Autopod_Skeleton_Figures/human_early_RNA_cluster_0.png") %>%
  image_ggplot()
plt1 <- image_read("~/Desktop/Autopod_Skeleton_Figures/human_early_RNA_cluster_1.png") %>%
  image_ggplot()
plt2 <- image_read("~/Desktop/Autopod_Skeleton_Figures/human_early_RNA_cluster_2.png") %>%
  image_ggplot()
plt3 <- image_read("~/Desktop/Autopod_Skeleton_Figures/human_early_RNA_cluster_3.png") %>%
  image_ggplot()
plt4 <- image_read("~/Desktop/Autopod_Skeleton_Figures/human_early_RNA_cluster_4.png") %>%
  image_ggplot()

#generate and save plot
human_RNA_clusters_wgcna <- plot_grid(plt1, plt2, plt3, plt4,  nrow =2, ncol = 2, labels = 1:4, align="hv")
ggsave(human_RNA_clusters_wgcna, device = "png", filename = "human_early_RNA_clusters_wgcna.png", path = "~/Desktop/Autopod_Skeleton_Figures/", width = 10, height = 10, dpi = 1000)
```

#late

```{r}
#glm design
# Again, we can include more things beside tissue, like we could look at the effect of tissue *and* age, or we could control for sex here. 
design <- formula(~replicate + limbtype)
#build DESeqDataSet
rsem.in <- DESeqDataSetFromMatrix(countData = ct_late, colData = sampleMetaData %>% dplyr::filter(timepoint == "Late"), design = design, tidy = T)

rsem_cols <-as.data.frame(row.names(rsem.in))
rsem_cols$ensembl <- sapply(strsplit(rsem_cols$`row.names(rsem.in)`, ".", fixed=T), function(x) x[1])

colnames(rsem_cols) <- c('ensembl_gene_id', 'ensembl')
rsem_annot <- merge(x= rsem_cols, y =  annot, by.x ='ensembl',
  by.y = 'ensembl_gene_id', no.dups = FALSE, all.x = TRUE, all.y=FALSE)

#combine gene symbols for duplicated rows
for (dup in rsem_annot$ensembl_gene_id[which(duplicated(rsem_annot$ensembl_gene_id))]){
  dup_ids <- which(rsem_annot$ensembl_gene_id == dup)
  dup_cat_id <- paste(rsem_annot$ensembl_gene_id[1], ", ", rsem_annot$ensembl_gene_id[2], sep ="")
  rsem_annot$hgnc_symbol[dup_ids[2]]<- dup_cat_id
}
rsem_annot_2 <- rsem_annot[-which(duplicated(rsem_annot$ensembl_gene_id)), ] %>% setorder(cols = "ensembl_gene_id")   
mcols(rsem.in) <- cbind(mcols(rsem.in), rsem_annot_2$hgnc_symbol)

#, test = "LRT", reduced = reducedest--pairwise does not need it. Or post-hoc tests can compare between pairs. 
rsem.de <- DESeq(rsem.in)
mcols(rsem.de) <- cbind(mcols(rsem.de), rsem_annot_2)

#plots related to dispersion, built in function to DESEQ
rsem.vst_late <- vst(rsem.de)

# Retrieve the normalized data from the `DESeqDataSet`
mat <- assay(rsem.vst_late)
mm <- model.matrix(~replicate, colData(rsem.vst_late))
mat <- limma::removeBatchEffect(mat, batch=colData(rsem.vst_late)$replicate)
assay(rsem.vst_late) <- mat
normalized_counts_late <- assay(rsem.vst_late)

#filter by DEGs
human_DEGs_late_list <- unique(human_DEGs_late$ensembl_gene_id)
DEG_normalized_counts_late <- normalized_counts_late[which(row.names(normalized_counts_late) %in% human_DEGs_late$ensembl_gene_id),]

#library(WGCNA)
input_mat = t(DEG_normalized_counts_late)

# Choose a set of soft-thresholding powers
powers = c(c(1:10), seq(from = 12, to = 30, by = 2))

# Call the network topology analysis function
sft = pickSoftThreshold(
  input_mat,             # <= Input data
  #blockSize = 30,
  powerVector = powers,
  verbose = 5
  )

par(mfrow = c(1,2));
cex1 = 0.9;

plot(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     xlab = "Soft Threshold (power)",
     ylab = "Scale Free Topology Model Fit, signed R^2",
     main = paste("Scale independence")
)
text(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     labels = powers, cex = cex1, col = "red"
)
abline(h = 0.90, col = "red")
plot(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     xlab = "Soft Threshold (power)",
     ylab = "Mean Connectivity",
     type = "n",
     main = paste("Mean connectivity")
)
text(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     labels = powers,
     cex = cex1, col = "red")

picked_power = min(which(sft$fitIndices$SFT.R.sq > 0.9))
temp_cor <- cor       
cor <- WGCNA::cor         # Force it to use WGCNA cor function (fix a namespace conflict issue)
netwk <- blockwiseModules(input_mat,                # <= input here

                          # == Adjacency Function ==
                          power = picked_power,                # <= power here
                          networkType = "signed",

                          # == Tree and Block Options ==
                          deepSplit = 2,
                          pamRespectsDendro = F,
                          # detectCutHeight = 0.75,
                          minModuleSize = 30,
                          maxBlockSize = 4000,

                          # == Module Adjustments ==
                          reassignThreshold = 0,
                          mergeCutHeight = 0.25,

                          # == TOM == Archive the run results in TOM file (saves time)
                          saveTOMs = T,
                          saveTOMFileBase = "ER",

                          # == Output Options
                          numericLabels = T,
                          verbose = 3)

module_df_late <- data.frame(
  gene_id = names(netwk$colors),
  cluster = netwk$colors,
  color = labels2colors(netwk$colors)
)

# Get Module Eigengenes per cluster
MEs <- moduleEigengenes(input_mat, netwk$colors)$eigengenes
MEs0 <- MEs

# Reorder modules so similar modules are next to each other
#MEs0 <- orderMEs(MEs0)
module_order = names(MEs0) %>% gsub("ME","", .)

# Add treatment names
MEs0$treatment <- sampleMetaData %>% dplyr::filter(timepoint == "Late") %>% pull(sampleType)

# tidy & plot data
mME = MEs0 %>%
  pivot_longer(-treatment) %>%
  mutate(
    color = gsub("ME", "", name),
    color = factor(color, levels = module_order) 
  ) %>% 
  group_by(treatment, color) %>% 
  summarize(avg_val = mean(value))
  
#create key to convert between cluster color and number
#Note: The grey color/module 0 is reserved for unassigned genes
color_cluster_key <- module_df_late %>% 
  group_by(cluster) %>% 
  summarize(gene_n = n()) 

write_delim(module_df_late %>% left_join(color_cluster_key) %>% 
  left_join(annot, by = join_by(gene_id == ensembl_gene_id)) %>% dplyr::select(gene_id, hgnc_symbol, cluster) %>% dplyr::rename(ensembl_gene_id= gene_id) %>% arrange(cluster),
            file = "~/Desktop/Capellini_Lab/Human Autopod RNA/human_late_gene_modules.tsv",
            delim = "\t")

for(i in min(color_cluster_key$cluster):max(color_cluster_key$cluster)){

  mME_temp <- mME %>% mutate(cluster = as.numeric(unfactor(color))) %>% 
    left_join(color_cluster_key) %>% 
    dplyr::filter(cluster == i) %>% 
                          mutate(value = avg_val) %>% 
                          mutate(tissue = treatment)
  
plot_autopod_expression(mME_temp %>% 
                          dplyr::select(tissue, value), 
                        file_name = paste("human_late_RNA_cluster_", i, sep = ""), opt_text = paste(mME_temp$gene_n[i+1], " genes", sep =""),
                        center =T
                        )
}

plt0 <- image_read("~/Desktop/Autopod_Skeleton_Figures/human_late_RNA_cluster_0.png") %>%
  image_ggplot()
plt1 <- image_read("~/Desktop/Autopod_Skeleton_Figures/human_late_RNA_cluster_1.png") %>%
  image_ggplot()
plt2 <- image_read("~/Desktop/Autopod_Skeleton_Figures/human_late_RNA_cluster_2.png") %>%
  image_ggplot()
plt3 <- image_read("~/Desktop/Autopod_Skeleton_Figures/human_late_RNA_cluster_3.png") %>%
  image_ggplot()
plt4 <- image_read("~/Desktop/Autopod_Skeleton_Figures/human_late_RNA_cluster_4.png") %>%
  image_ggplot()
plt5 <- image_read("~/Desktop/Autopod_Skeleton_Figures/human_late_RNA_cluster_5.png") %>%
  image_ggplot()
plt6 <- image_read("~/Desktop/Autopod_Skeleton_Figures/human_late_RNA_cluster_6.png") %>%
  image_ggplot()
plt7 <- image_read("~/Desktop/Autopod_Skeleton_Figures/human_late_RNA_cluster_7.png") %>%
  image_ggplot()
plt8 <- image_read("~/Desktop/Autopod_Skeleton_Figures/human_late_RNA_cluster_8.png") %>%
  image_ggplot()
plt9 <- image_read("~/Desktop/Autopod_Skeleton_Figures/human_late_RNA_cluster_9.png") %>%
  image_ggplot()
plt10 <- image_read("~/Desktop/Autopod_Skeleton_Figures/human_late_RNA_cluster_10.png") %>%
  image_ggplot()
plt11 <- image_read("~/Desktop/Autopod_Skeleton_Figures/human_late_RNA_cluster_11.png") %>%
  image_ggplot()
plt12 <- image_read("~/Desktop/Autopod_Skeleton_Figures/human_late_RNA_cluster_12.png") %>%
  image_ggplot()
plt13 <- image_read("~/Desktop/Autopod_Skeleton_Figures/human_late_RNA_cluster_13.png") %>%
  image_ggplot()
plt14 <- image_read("~/Desktop/Autopod_Skeleton_Figures/human_late_RNA_cluster_14.png") %>%
  image_ggplot()

#generate and save plot
human_RNA_clusters_wgcna <- plot_grid(plt1, plt2, plt3, plt4, plt5, plt6, plt7, plt8, plt9, plt10, plt11, plt12, plt13, plt14, nrow =3, ncol = 5, labels = 1:14, align="hv")
ggsave(human_RNA_clusters_wgcna, device = "png", filename = "human_late_RNA_clusters_wgcna.png", path = "~/Desktop/Autopod_Skeleton_Figures/", width = 10, height = 10, dpi = 1000)
```

# GO Enrichments

```{r}
# note the DESeq2 cannot handle perfectly correlated variables like individual and sex since one contains a full set of the other
# due to this, I only control for individual (replicate)
design <- formula(~replicate + tissue)

#build DESeqDataSet
human_rsem.in <- DESeqDataSetFromMatrix(countData = ct, colData = sampleMetaData, design = design, tidy = T)
#, test = "LRT", reduced = reducedest--pairwise does not need it. Or post-hoc tests can compare between pairs. 
human_rsem.de <- DESeq(human_rsem.in)

#run combined analysis to use GO enrichment code
# Retrieve the normalized data from the `DESeqDataSet`
# do not correct for replicate when comparing timepoints, this removes timepoint effects
# run the PCA DESeq to generate the human_rsem.de object
human_rsem.vst <- vst(human_rsem.de)
normalized_counts <- assay(human_rsem.vst)

#filter by DEGs
human_DEGs_list <- unique(human_DEGs2$ensembl_gene_id)
DEG_normalized_counts <- normalized_counts[which(row.names(normalized_counts) %in% human_DEGs_list),]

#library(WGCNA)
input_mat = t(DEG_normalized_counts)

# Choose a set of soft-thresholding powers
powers = c(c(1:10), seq(from = 12, to = 30, by = 2))

# Call the network topology analysis function
sft = pickSoftThreshold(
  input_mat,             # <= Input data
  #blockSize = 30,
  powerVector = powers,
  verbose = 5
  )

par(mfrow = c(1,2));
cex1 = 0.9;

plot(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     xlab = "Soft Threshold (power)",
     ylab = "Scale Free Topology Model Fit, signed R^2",
     main = paste("Scale independence")
)
text(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     labels = powers, cex = cex1, col = "red"
)
abline(h = 0.90, col = "red")
plot(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     xlab = "Soft Threshold (power)",
     ylab = "Mean Connectivity",
     type = "n",
     main = paste("Mean connectivity")
)
text(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     labels = powers,
     cex = cex1, col = "red")

#choose power as minimum value to produce Rsq greater than 0.9
picked_power = min(which(sft$fitIndices$SFT.R.sq > 0.9))
#picking first value below 0.9 before R.sq declines
picked_power = 8
temp_cor <- cor       
cor <- WGCNA::cor         # Force it to use WGCNA cor function (fix a namespace conflict issue)
netwk <- blockwiseModules(input_mat,                # <= input here

                          # == Adjacency Function ==
                          power = picked_power,                # <= power here
                          networkType = "signed",

                          # == Tree and Block Options ==
                          deepSplit = 2,
                          pamRespectsDendro = F,
                          # detectCutHeight = 0.75,
                          minModuleSize = 30,
                          maxBlockSize = 4000,

                          # == Module Adjustments ==
                          reassignThreshold = 0,
                          mergeCutHeight = 0.25,

                          # == TOM == Archive the run results in TOM file (saves time)
                          saveTOMs = T,
                          saveTOMFileBase = "ER",

                          # == Output Options
                          numericLabels = T,
                          verbose = 3)

module_df <- data.frame(
  gene_id = names(netwk$colors),
  cluster = netwk$colors,
  color = labels2colors(netwk$colors)
)

#run gene ontology analysis
cluster_go_analysis(cluster_results = module_df, 
                    universe = ct$ensembl_gene_id, 
                    database = org.Hs.eg.db, 
                    outfile = "~/Desktop/Capellini_Lab/Human Autopod RNA/Human_RNA_Clusters_Go_Enrichments.csv")

#read in enrichments 
human_GO_clusters <- unique(read.delim(file= "~/Desktop/Capellini_Lab/Human Autopod RNA/Human_RNA_Clusters_Go_Enrichments.csv", sep = ",", header = T))

#run eary stage gene ontology analysis
cluster_go_analysis(cluster_results = module_df_early, 
                    universe = ct$ensembl_gene_id, 
                    database = org.Hs.eg.db, 
                    outfile = "~/Desktop/Capellini_Lab/Human Autopod RNA/Human_RNA_Clusters_Early_GO_Enrichments.csv")

#read in enrichments 
human_GO_clusters_early <- unique(read.delim(file= "~/Desktop/Capellini_Lab/Human Autopod RNA/Human_RNA_Clusters_Early_GO_Enrichments.csv", sep = "\t", header = T))

#run late stage gene ontology analysis
cluster_go_analysis(cluster_results = module_df_late, 
                    universe = ct$ensembl_gene_id, 
                    database = org.Hs.eg.db, 
                    outfile = "~/Desktop/Capellini_Lab/Human Autopod RNA/Human_RNA_Clusters_Late_GO_Enrichments.csv")

#read in enrichments 
human_GO_clusters_late <- unique(read.delim(file= "~/Desktop/Capellini_Lab/Human Autopod RNA/Human_RNA_Clusters_Late_GO_Enrichments.csv", sep = "\t", header = T))

#create normalized counts for function
norm_counts <- normalized_counts %>% as.data.frame() %>% mutate(ensembl_gene_id = row.names(normalized_counts)) %>% pivot_longer(!ensembl_gene_id, names_to = "name", values_to = "count") %>% right_join(y = annot[,1:2]) 
norm_counts$name <- gsub(pattern = "E", replacement = "", x = norm_counts$name)
norm_counts$name <- gsub(pattern = "L", replacement = "", x = norm_counts$name)

write_tsv(norm_counts, file = "~/Desktop/Capellini_Lab/Human Autopod RNA/human_normalized_counts.tsv")

```

#repeat key steps for just early samples 1,3,5 to compare with mouse

```{r}
sampleTypes135 <- paste(unique(sampleMetaData$sampleType)[c(1,3,5:6, 8, 10:11, 13, 15:16, 18, 20)], "E", sep = "")

sample_unique_info135 <- data.frame(sample = sampleTypes135, comparisons1 = NA, comparisons2= NA)

#create sample and control sets
for (samp in sampleTypes135){
  comp_digits <- list()
  alt_region <- "Unknown"
  digit_comp <- "Unknown"
  neighbors <- "Unknown"
  samp_info <-unlist(strsplit(samp, ""))
  #check if digit 1
  if (samp_info[3] == "1"){
    comp_digits <- 3
  }
  #check if digit 3
  if (samp_info[3] == "3"){
    comp_digits <- c(1, 5)
  }
  #check if digit 5
  if (samp_info[3] == "5"){
    comp_digits <- c(3)
  }
  #Get other region
  #check if metapodial
  if (samp_info[2] == "M"){
    alt_region <- "P"
  }
  #check if phalanges
  if (samp_info[2] == "P"){
    alt_region <- "M"
  }
  #check if hand (forelimb)
  if (samp_info[1] == "H"){
    alt_limb <- "F"
  }
  #check if phalanges
  if (samp_info[1] == "F"){
    alt_limb <- "H"
  }
        digit_comp <- paste(samp_info[1], alt_region, samp_info[3], samp_info[4], sep = "")
    neighbors <- list()
    for (n in 1:length(comp_digits)){
      neighbors <- append(neighbors, paste(samp_info[1], samp_info[2], comp_digits[n], samp_info[4], sep = ""))
    }
  neighbors2 <- paste(neighbors, collapse ="|")
  limb_comp <- paste(alt_limb, samp_info[2], samp_info[3], samp_info[4], sep = "")
  sample_unique_info135$comparisons1[which(sample_unique_info135$sample == samp) ] <- list(as.list(c(neighbors, digit_comp, limb_comp)))
  sample_unique_info135$comparisons2[which(sample_unique_info135$sample == samp) ] <- paste(samp, neighbors2, digit_comp, limb_comp, sep = "|")
}

#add key to convert between my nomenclature and something less easy to misinterpret
autopod_key_notp <- data.frame(autopod_abrevs = c("FM1", "FM2", "FM3", "FM4", "FM5", 
                    "FP1", "FP2", "FP3", "FP4", "FP5", 
                    "HM1", "HM2", "HM3", "HM4", "HM5", 
                    "HP1", "HP2", "HP3", "HP4", "HP5"), autopod_tissues = c("METATARSAL1", "METATARSAL2", "METATARSAL3", "METATARSAL4", "METATARSAL5", 
                                   "FOOTPHALANGES1", "FOOTPHALANGES2", "FOOTPHALANGES3", "FOOTPHALANGES4", "FOOTPHALANGES5", 
                                   "METACARPAL1", "METACARPAL2", "METACARPAL3", "METACARPAL4", "METACARPAL5", 
                                   "HANDPHALANGES1", "HANDPHALANGES2", "HANDPHALANGES3", "HANDPHALANGES4", "HANDPHALANGES5"))
autopod_key_early <- autopod_key_notp 
autopod_key_late <- autopod_key_notp 
autopod_key_early$autopod_abrevs <- paste(autopod_key_early$autopod_abrevs, "E", sep = "")
autopod_key_late$autopod_abrevs <- paste(autopod_key_late$autopod_abrevs, "L", sep = "")
autopod_key_early$autopod_tissues <- paste(autopod_key_early$autopod_tissues, "E", sep = "")
autopod_key_late$autopod_tissues <- paste(autopod_key_late$autopod_tissues, "L", sep = "")
  
autopod_key <- rbind(autopod_key_early, autopod_key_late)
autopod_key135 <- autopod_key_early[c(1,3,5:6, 8, 10:11, 13, 15:16, 18, 20),]

# #create dataframe to store results
 human_DEGs135 <- data.frame()
#
 #iterate over each sample
 for (i in 1:length(sampleTypes135)){
   #find number of
   comp_samples <- unlist(sample_unique_info135$comparisons1[i])
   for(c in 1:length(comp_samples)){
     #filter data and metadata
     colsToKeep <- c(1, grep(paste(sample_unique_info135$sample[i], comp_samples[c], sep = "|"), names(ct)))
     good_data <- ct[,colsToKeep]
     filteredMetaData <- sampleMetaData %>% dplyr::filter(sampleTypeTP %in% c(sample_unique_info135$sample[i], comp_samples[c]))
filteredMetaData$sampleTypeTP <- factor(filteredMetaData$sampleTypeTP, levels = c(sample_unique_info135$sample[i], comp_samples[c]))

     test_output <- runDESEQ2_df_hs(design_in = formula(~replicate + sampleTypeTP), count_data = good_data, col_data = filteredMetaData) %>% setDT(keep.rownames = T)
     #fix column names
     names(test_output)[1] <- "ensembl_gene_id"
     names(test_output)[8] <- "hgnc_symbol"
     test_output$tissue1 <- autopod_key135$autopod_tissues[which(autopod_key135$autopod_abrevs == sample_unique_info135$sample[i])]
     test_output$tissue2 <- autopod_key135$autopod_tissues[which(autopod_key135$autopod_abrevs == comp_samples[c])]
     if(nrow(test_output) > 0){
     human_DEGs <- rbind(human_DEGs135, test_output)
     }
     #note if no DEGs identified
     if(nrow(test_output) == 0){
     test_output <- data.frame(ensembl_gene_id = NA,  baseMean = NA, log2FoldChange = NA, lfcSE = NA, stat = NA, pvalue= NA, padj = NA, hgnc_symbol= "NO_DEGs_FOUND", tissue1=autopod_key135$autopod_tissues[which(autopod_key135$autopod_abrevs == sample_unique_info135$sample[i])], tissue2= autopod_key135$autopod_tissues[which(autopod_key135$autopod_abrevs == comp_samples[c])])
     }
     human_DEGs135 <- rbind(human_DEGs135, test_output)
   }
 }
 
 sampleMetaData_early135 <- sampleMetaData %>% dplyr::filter(timepoint == "Early") %>% dplyr::filter(digit %in% c("I", "III", "V"))
 #now identify gradient genes
dds <- DESeqDataSetFromMatrix(countData = ct[, c(1, which(colnames(ct) %in% rownames(sampleMetaData_early135)))], colData = sampleMetaData_early135, design = ~replicate + digit, tidy = T)
dds <- DESeq(dds, test="LRT", reduced=~replicate)
dds_res <- results(dds, alpha = 0.05)
dds_res_df <- merge(x= as.data.frame(dds_res), by.x ='row.names',  y = annot,
  by.y = 'ensembl_gene_id', no.dups = TRUE) %>% dplyr::filter(padj < 0.05)

#reform to add to human DEG object
colnames(dds_res_df)[1] <- "ensembl_gene_id"
dds_res_df$tissue1 <- "GRADIENT"
dds_res_df$tissue2 <- "GRADIENT"
dds_res_df$gene_biotype <- NULL

human_DEGs1352 <- rbind(human_DEGs135, dds_res_df) %>% dplyr::filter(padj < 0.05, abs(log2FoldChange) > 0.58 )

write_delim(x = as.data.frame(human_DEGs1352), file= "/Users/alexanderokamoto/Desktop/Capellini_Lab/Human Autopod RNA/All_Human_Bio_Pairwise_Autopod_DEGs135_early.tsv", delim = "\t")

human_DEGs135 <- unique(read.delim(file= "/Users/alexanderokamoto/Desktop/Capellini_Lab/Human Autopod RNA/All_Human_Bio_Pairwise_Autopod_DEGs135_late.tsv", sep = "\t", header = T))

length(intersect(human_DEGs_early$ensembl_gene_id, human_DEGs1352$ensembl_gene_id))
```

## 135 comparison late

```{r}
autopod_key135 <- autopod_key_late[c(1,3,5:6, 8, 10:11, 13, 15:16, 18, 20),]
sampleTypes135_late <- gsub(pattern = "E", x = sampleTypes135, replacement = "L", )

sampleTypes135 <- paste(unique(sampleMetaData$sampleType)[c(1,3,5:6, 8, 10:11, 13, 15:16, 18, 20)], "L", sep = "")

sample_unique_info135 <- data.frame(sample = sampleTypes135, comparisons1 = NA, comparisons2= NA)

#create sample and control sets
for (samp in sampleTypes135){
  comp_digits <- list()
  alt_region <- "Unknown"
  digit_comp <- "Unknown"
  neighbors <- "Unknown"
  samp_info <-unlist(strsplit(samp, ""))
  #check if digit 1
  if (samp_info[3] == "1"){
    comp_digits <- 3
  }
  #check if digit 3
  if (samp_info[3] == "3"){
    comp_digits <- c(1, 5)
  }
  #check if digit 5
  if (samp_info[3] == "5"){
    comp_digits <- c(3)
  }
  #Get other region
  #check if metapodial
  if (samp_info[2] == "M"){
    alt_region <- "P"
  }
  #check if phalanges
  if (samp_info[2] == "P"){
    alt_region <- "M"
  }
  #check if hand (forelimb)
  if (samp_info[1] == "H"){
    alt_limb <- "F"
  }
  #check if phalanges
  if (samp_info[1] == "F"){
    alt_limb <- "H"
  }
        digit_comp <- paste(samp_info[1], alt_region, samp_info[3], samp_info[4], sep = "")
    neighbors <- list()
    for (n in 1:length(comp_digits)){
      neighbors <- append(neighbors, paste(samp_info[1], samp_info[2], comp_digits[n], samp_info[4], sep = ""))
    }
  neighbors2 <- paste(neighbors, collapse ="|")
  limb_comp <- paste(alt_limb, samp_info[2], samp_info[3], samp_info[4], sep = "")
  sample_unique_info135$comparisons1[which(sample_unique_info135$sample == samp) ] <- list(as.list(c(neighbors, digit_comp, limb_comp)))
  sample_unique_info135$comparisons2[which(sample_unique_info135$sample == samp) ] <- paste(samp, neighbors2, digit_comp, limb_comp, sep = "|")
}

# #create dataframe to store results
 human_DEGs135 <- data.frame()
#
 #iterate over each sample
 for (i in 1:length(sampleTypes135_late)){
   #find number of
   comp_samples <- unlist(sample_unique_info135$comparisons1[i])
   for(c in 1:length(comp_samples)){
     #filter data and metadata
     colsToKeep <- c(1, grep(paste(sample_unique_info135$sample[i], comp_samples[c], sep = "|"), names(ct)))
     good_data <- ct[,colsToKeep]
     filteredMetaData <- sampleMetaData %>% dplyr::filter(sampleTypeTP %in% c(sample_unique_info135$sample[i], comp_samples[c]))
filteredMetaData$sampleTypeTP <- factor(filteredMetaData$sampleTypeTP, levels = c(sample_unique_info135$sample[i], comp_samples[c]))

     test_output <- runDESEQ2_df_hs(design_in = formula(~replicate + sampleTypeTP), count_data = good_data, col_data = filteredMetaData) %>% setDT(keep.rownames = T)
     #fix column names
     names(test_output)[1] <- "ensembl_gene_id"
     names(test_output)[8] <- "hgnc_symbol"
     test_output$tissue1 <- autopod_key135$autopod_tissues[which(autopod_key135$autopod_abrevs == sample_unique_info135$sample[i])]
     test_output$tissue2 <- autopod_key135$autopod_tissues[which(autopod_key135$autopod_abrevs == comp_samples[c])]
     if(nrow(test_output) > 0){
     human_DEGs <- rbind(human_DEGs135, test_output)
     }
     # #note if no DEGs identified
     # if(nrow(test_output) == 0){
     # test_output <- data.frame(ensembl_gene_id = NA,  baseMean = NA, log2FoldChange = NA, lfcSE = NA, stat = NA, pvalue= NA, padj = NA, hgnc_symbol= "NO_DEGs_FOUND", tissue1=autopod_key135$autopod_tissues[which(autopod_key135$autopod_abrevs == sample_unique_info135$sample[i])], tissue2= autopod_key135$autopod_tissues[which(autopod_key135$autopod_abrevs == comp_samples[c])])
     # }
     if(nrow(test_output) > 0){
      human_DEGs135 <- rbind(human_DEGs135, test_output)
     }
   }
 }
 
 sampleMetaData_late135 <- sampleMetaData %>% dplyr::filter(timepoint == "Late") %>% dplyr::filter(digit %in% c("I", "III", "V"))
 #now identify gradient genes
dds <- DESeqDataSetFromMatrix(countData = ct[, c(1, which(colnames(ct) %in% rownames(sampleMetaData_late135)))], colData = sampleMetaData_late135, design = ~replicate + digit, tidy = T)
dds <- DESeq(dds, test="LRT", reduced=~replicate)
dds_res <- results(dds, alpha = 0.05)
dds_res_df <- merge(x= as.data.frame(dds_res), by.x ='row.names',  y = annot,
  by.y = 'ensembl_gene_id', no.dups = TRUE) %>% dplyr::filter(padj < 0.05)

#reform to add to human DEG object
colnames(dds_res_df)[1] <- "ensembl_gene_id"
dds_res_df$tissue1 <- "GRADIENT"
dds_res_df$tissue2 <- "GRADIENT"
dds_res_df$gene_biotype <- NULL

human_DEGs1352 <- rbind(human_DEGs135, dds_res_df) %>% dplyr::filter(padj < 0.05, abs(log2FoldChange) > 0.58 )

write_delim(x = as.data.frame(human_DEGs1352), file= "/Users/alexanderokamoto/Desktop/Capellini_Lab/Human Autopod RNA/All_Human_Bio_Pairwise_Autopod_DEGs135_late.tsv", delim = "\t")

human_DEGs135_late <- unique(read.delim(file= "/Users/alexanderokamoto/Desktop/Capellini_Lab/Human Autopod RNA/All_Human_Bio_Pairwise_Autopod_DEGs135_late.tsv", sep = "\t", header = T))

length(intersect(human_DEGs_late$ensembl_gene_id, human_DEGs135_late$ensembl_gene_id))
```

##comparison of human autopod module sets

```{r}
#load data
human_modules <- read_tsv(file = "~/Desktop/Capellini_Lab/Human Autopod RNA/human_gene_clusters_both_tps.tsv")
human_modules_early <- read_tsv(file = "~/Desktop/Capellini_Lab/Human Autopod RNA/human_early_gene_modules.tsv")
human_modules_late <- read_tsv(file = "~/Desktop/Capellini_Lab/Human Autopod RNA/human_late_gene_modules.tsv")
mouse_modules <- read_tsv(file = "~/Desktop/Capellini_Lab/Mouse Autopod RNA/mouse_gene_clusters.tsv")

length(unique(human_modules$cluster))
length(unique(human_modules_early$cluster))
length(unique(human_modules_late$cluster))

for (i in 1:12){
  shared_genes <- length(intersect(human_modules_late$ensembl_gene_id[which(human_modules_late$cluster == 5)], human_modules$ensembl_gene_id[which(human_modules$cluster == i)]))
 print(paste(shared_genes, " shared genes with module ", i, sep = "")) 
}

for (i in 1:12){
  shared_genes <- length(intersect(human_modules_early$ensembl_gene_id[which(human_modules_early$cluster == 2)], human_modules$ensembl_gene_id[which(human_modules$cluster == i)]))
 print(paste(shared_genes, " shared genes with module ", i, sep = "")) 
}

for (i in 1:12){
  shared_genes <- length(intersect(human_modules_early$ensembl_gene_id[which(human_modules_early$cluster == 4)], human_modules_late$ensembl_gene_id[which(human_modules_late$cluster == i)]))
 print(paste(shared_genes, " shared genes with module ", i, sep = "")) 
}

for (i in 1:12){
  shared_genes <- length(intersect(human_modules$ensembl_gene_id[which(human_modules$cluster == 5)], mouse_to_human_gene(mouse_modules$ensembl_gene_id[which(mouse_modules$cluster == i)])))
 print(paste(shared_genes, " shared genes with module ", i, sep = "")) 
}

#DEGs per comparison
mouse_DEGs <- unique(read.delim(file= "/Users/alexanderokamoto/Desktop/Capellini_Lab/Mouse Autopod RNA/All_Mouse_Bio_Pairwise_Autopod_DEGs.tsv", sep = "\t", header = T))

length(intersect(human_DEGs2$ensembl_gene_id, mouse_to_human_gene(mouse_DEGs$ensembl_gene_id)))
```

##timepoint analysis

```{r}
timepoint_DEGs <- runDESEQ2_df_hs(design_in = formula(~timepoint), count_data = ct, col_data =  sampleMetaData) %>% dplyr::filter(abs(log2FoldChange) > 0.58) %>% dplyr::filter(padj < 0.05)

length(which(timepoint_DEGs$log2FoldChange > 0))
length(which(timepoint_DEGs$log2FoldChange < 0))
timepoint_DEGs$timepoint<- "EARLY"
timepoint_DEGs$timepoint[which(timepoint_DEGs$log2FoldChange > 0)] <- "LATE"

#timepoint_DEGs <- read.delim(file= "/Users/alexanderokamoto/Desktop/Capellini_Lab/Human Autopod RNA/Human_Autopod_Timepoint_DEGs.tsv", sep = "\t", header = T)
write.table(x = timepoint_DEGs , file = "/Users/alexanderokamoto/Desktop/Capellini_Lab/Human Autopod RNA/Human_Autopod_Timepoint_DEGs.tsv", sep = "\t", col.names = T)

late_GO <- simple_go_analysis(timepoint_DEGs$rsem_annot.hgnc_symbol[which(timepoint_DEGs$log2FoldChange > 0)])
late_GO$timepoint <- "LATE"

early_GO <- simple_go_analysis(timepoint_DEGs$rsem_annot.hgnc_symbol[which(timepoint_DEGs$log2FoldChange < 0)])
early_GO$timepoint <- "EARLY"

timepoint_GO <- rbind(early_GO, late_GO)

  timepoint_GO$GeneRatio <- paste("'", timepoint_GO$GeneRatio, "'", sep ="")
  timepoint_GO$BgRatio <- paste("'", timepoint_GO$BgRatio, "'", sep ="")
  write.csv(timepoint_GO, file = "/Users/alexanderokamoto/Desktop/Capellini_Lab/Human Autopod RNA/Human_Autopod_Timepoint_DEGs_GO_enrichments.csv", row.names = F)
```
