---
title: "Human_Mouse_ATAC_Data_Cluster_Processing"
author: "Alexander Okamoto"
date: "2025-12-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Human ATAC-seq Data Analysis

# Initially processing of ATAC-seq sequencing data was performed on the Harvard FAS Cluster. Examples of the code run are included below. 

#Steps included:

## Adapter Trimming

To trim adapaters I used NGmerge (Gaspar et al. *Bioinformatics* 2018). The -e 30 parameter was chosen based on <https:/github.com/jsh58/NGmerge/issues/6>.


##Alignment to the genome

After adapter trimming, reads were aligning to the human genome. To do this, I downloaded the UCSC hg38 genome prebuild reference from Illumina (http://igenomes.illumina.com.s3-website-us-east-1.amazonaws.com/Homo_sapiens/UCSC/hg38/Homo_sapiens_UCSC_hg38.tar.gz)

##Duplicate/mitochrondrial read removal
I used Picard to remove duplicate reads, then John Gaspar’s removeChrom.pymethod to remove mitochondrial reads.

## Peak Calling using MACS2

## Identify reproducible consensus peaks using IDR.

I then use Daniel Richard’s IDR pipeline to call consensus peaks, which is contained in the file "IDR_script_APRIL2021_CUSTOM_CUTOFFS.R"

## Here is the script that combines all this to fully process the tissues from a single individual sample to MACS2 peak calling

```{bash, eval=FALSE}

#load modules
module load NGmerge/0.2-fasrc01
module load centos6/0.0.1-fasrc01
module load intel/2017c impi/2017.4.239 Bowtie2/2.3.4.1
module load samtools/1.10-fasrc01
module load picard/2.9.0-fasrc01
module load ATAC-seq
module load macs2/2.1.1.20160309-fasrc02

#set variables
rep_id="28829"
location1="/n/capellini_lab/users/aokamoto/Autopod_Project/Sequencing_Data/221129_A00794_0742_AHGWKNDSX5_SUB13550/fastq/H28829_other"
location2="/n/capellini_lab/users/aokamoto/Autopod_Project/Sequencing_Data/221129_A00794_0742_AHGWKNDSX5_SUB13550/fastq/H28829_mouse2"

#if needed, make directory for output reads

if [ -d "/n/holyscratch01/capellini_lab/aokamoto/Human_cat_ATAC_files" ]; 
then
  echo "Output directory /n/holyscratch01/capellini_lab/aokamoto/Human_cat_ATAC_files exists."
  
else
  mkdir /n/holyscratch01/capellini_lab/aokamoto/Human_cat_ATAC_files
fi

#if needed, make directory for trimmed reads

if [ -d "/n/holyscratch01/capellini_lab/aokamoto/Human_Trimmed_Reads" ]; 
then
  echo "Output directory /n/holyscratch01/capellini_lab/aokamoto/Human_Trimmed_Reads exists."
  
else
  mkdir /n/holyscratch01/capellini_lab/aokamoto/Human_Trimmed_Reads
fi

#if needed, make directory for aligned reads
if [ -d "/n/holyscratch01/capellini_lab/aokamoto/Human_Aligned" ]; 
then
  echo "Output directory /n/holyscratch01/capellini_lab/aokamoto/Human_Aligned exists."
  
else
  mkdir /n/holyscratch01/capellini_lab/aokamoto/Human_Aligned
fi

#if needed, make directory for Picard output
if [ -d "/n/holyscratch01/capellini_lab/aokamoto/Human_Picard_output" ]; 
then
  echo "Output directory /n/holyscratch01/capellini_lab/aokamoto/Human_Picard_output/ exists."
  
else
  mkdir /n/holyscratch01/capellini_lab/aokamoto/Human_Picard_output
fi

#if needed, make directory for Gaspar remove mitochondria output
if [ -d "/n/holyscratch01/capellini_lab/aokamoto/Human_noMT" ]; 
then
  echo "Output directory /n/holyscratch01/capellini_lab/aokamoto/Human_noMT exists."
  
else
  mkdir /n/holyscratch01/capellini_lab/aokamoto/Human_noMT
fi

#iterate over samples
for sample in HP1 HP2 HP3 HP4 HP5 HM1 HM2 HM3 HM4 HM5 FP1 FP2 FP3 FP4 FP5 FM1 FM2 FM3 FM4 FM5

do 

#concatenate reads if needed because a sample had to be run twice
file1=$(find ${location1} -name ${rep_id}_${sample}*_R1_001.fastq.gz)
file3=$(find ${location1} -name ${rep_id}_${sample}*_R2_001.fastq.gz)
file2=$(find ${location2} -name ${rep_id}_${sample}*_R1_001.fastq.gz)
file4=$(find ${location2} -name ${rep_id}_${sample}*_R2_001.fastq.gz)

#determine how many files found an concatenate as needed
if [[ -f ${file1} && ! -f ${file2} ]]; 
then
  cp ${file1} /n/holyscratch01/capellini_lab/aokamoto/Human_cat_ATAC_files/${rep_id}_${sample}_R1.fastq.gz
  cp ${file3} /n/holyscratch01/capellini_lab/aokamoto/Human_cat_ATAC_files/${rep_id}_${sample}_R2.fastq.gz
  echo "Only ${file1} found."
elif [[ ! -f ${file1} && -f ${file2} ]]; then
  cp ${file2} /n/holyscratch01/capellini_lab/aokamoto/Human_cat_ATAC_files/${rep_id}_${sample}_R1.fastq.gz
  cp ${file4} /n/holyscratch01/capellini_lab/aokamoto/Human_cat_ATAC_files/${rep_id}_${sample}_R2.fastq.gz
  echo "Only ${file2} found."
elif [[ -f ${file1} && -f ${file2} ]]; then
  cat ${file1} ${file2} > /n/holyscratch01/capellini_lab/aokamoto/Human_cat_ATAC_files/${rep_id}_${sample}_R1.fastq.gz
  cat ${file3} ${file4} > /n/holyscratch01/capellini_lab/aokamoto/Human_cat_ATAC_files/${rep_id}_${sample}_R2.fastq.gz
  echo "${file1} and ${file2} found."
fi

echo "Trimming adaptors using NGmerge"

NGmerge -a -1 /n/holyscratch01/capellini_lab/aokamoto/Human_cat_ATAC_files/${rep_id}_${sample}_R1.fastq.gz -2 /n/holyscratch01/capellini_lab/aokamoto/Human_cat_ATAC_files/${rep_id}_${sample}_R2.fastq.gz -o /n/holyscratch01/capellini_lab/aokamoto/Human_Trimmed_Reads/${rep_id}_${sample}_trimmed -v -e 30 -n 48

echo "Aligning ${sample} to UCSC hg38 genome."

bowtie2 \
-X 2000 \
-t \
-p 48 \
-x /n/capellini_lab/users/aokamoto/Autopod_Project/HumanReference/Homo_sapiens/UCSC/hg38/Sequence/Bowtie2Index/genome \
--very-sensitive \
-1 /n/holyscratch01/capellini_lab/aokamoto/Human_Trimmed_Reads/${rep_id}_${sample}_trimmed_1.fastq.gz \
-2 /n/holyscratch01/capellini_lab/aokamoto/Human_Trimmed_Reads/${rep_id}_${sample}_trimmed_2.fastq.gz | samtools view -bS - | samtools sort -o /n/holyscratch01/capellini_lab/aokamoto/Human_Aligned/${rep_id}_${sample}_coordSort.bam

echo "Align and sort for ${rep_id} ${sample} completed."

samtools index /n/holyscratch01/capellini_lab/aokamoto/Human_Aligned/${rep_id}_${sample}_coordSort.bam /n/holyscratch01/capellini_lab/aokamoto/Human_Aligned/${rep_id}_${sample}_coordSort.bai

echo "Indexing for ${rep_id} ${sample} completed."

sample_file="/n/holyscratch01/capellini_lab/aokamoto/Human_Aligned/${rep_id}_${sample}_coordSort.bam"
outfile="/n/holyscratch01/capellini_lab/aokamoto/Human_Picard_output/${rep_id}_${sample}_picard-mrkdup.bam"
metrics="/n/holyscratch01/capellini_lab/aokamoto/Human_Picard_output/${rep_id}_${sample}_picard-mrkdup_metrics.txt"

echo "Marking duplicates for ${rep_id} ${sample}"

java -jar $PICARD_TOOLS_HOME/picard.jar MarkDuplicates \
      I=$sample_file \
      O=$outfile \
      M=$metrics

#index new file with duplicates removed
samtools index $outfile

inBAM="/n/holyscratch01/capellini_lab/aokamoto/Human_Picard_output/${rep_id}_${sample}_picard-mrkdup.bam"
noMT="/n/holyscratch01/capellini_lab/aokamoto/Human_noMT/${rep_id}_${sample}.noMT.bam"

echo "Removing mito reads from ${inBAM} using removeChrom.py"

#output new bam file without mitochondrial reads
samtools view -h $inBAM | removeChrom - - chrM | samtools view -b - > $noMT

echo "Calling peaks for ${sample} using MACS2"

macs2 callpeak \
   -t ${noMT} \
   -n ${rep_id}_${sample} \
   -f BAMPE \
   -g hs \
   --outdir "/n/holyscratch01/capellini_lab/aokamoto/macs2output" \
   --nolambda \
   --keep-dup all \
   --bdg \
   --verbose 3

echo "Success!"
done
```

#Now, call IDR script

```{bash, eval = FALSE}
module load centos6/0.0.1-fasrc01
module load macs2/2.1.1.20160309-fasrc02
module load samtools/1.10-fasrc01
module load Anaconda3/2020.11
module load R/4.1.0-fasrc01
source activate /n/home04/aokamoto/miniconda3/envs/idr_py

id_1='29067'
id_2='29042'
id_3='28680'
stage="late"

for sample in HP1 HP2 HP4 HP5 HM1 HM2 HM4 HM5 FP1 FP2 FP3 FP4 FP5 FM1 FM2 FM3 FM4 FM5

do

inbam1="/n/holyscratch01/capellini_lab/aokamoto/Human_noMT/${id_1}_${sample}.noMT.bam"
inbam2="/n/holyscratch01/capellini_lab/aokamoto/Human_noMT/${id_2}_${sample}.noMT.bam"
inbam3="/n/holyscratch01/capellini_lab/aokamoto/Human_noMT/${id_3}_${sample}.noMT.bam"

inpeaks1="/n/holyscratch01/capellini_lab/aokamoto/macs2output/${id_1}_${sample}_peaks.narrowPeak"
inpeaks2="/n/holyscratch01/capellini_lab/aokamoto/macs2output/${id_2}_${sample}_peaks.narrowPeak"
inpeaks3="/n/holyscratch01/capellini_lab/aokamoto/macs2output/${id_3}_${sample}_peaks.narrowPeak"

echo "Creating combined peak file for ${sample}"
rm ${sample}_${stage}_peakset.txt
echo ${inpeaks1} > ${sample}_${stage}_peakset.txt
echo ${inpeaks2} >> ${sample}_${stage}_peakset.txt
echo ${inpeaks3} >> ${sample}_${stage}_peakset.txt

echo "Merging replicates for  ${sample} using samtools"

samtools merge -o "/n/holyscratch01/capellini_lab/aokamoto/Human_noMT/${sample}_${stage}_pooled.bam" ${inbam1} ${inbam2} ${inbam3}

echo "Calling peaks for ${sample} using MACS2"

macs2 callpeak \
   -t "/n/holyscratch01/capellini_lab/aokamoto/Human_noMT/${sample}_${stage}_pooled.bam" \
   -n ${sample}_${stage}_pooled \
   -f BAMPE \
   -g hs \
   --outdir "/n/holyscratch01/capellini_lab/aokamoto/macs2output" \
   --nolambda \
   --keep-dup all \
   --bdg \
   --verbose 3
   
echo "Calculating IDR peak set for ${stage} ${sample}"

Rscript IDR_script_APRIL2021_CUSTOM_CUTOFFS.R idr_calling_updated ${sample}_${stage}_peakset.txt /n/holyscratch01/capellini_lab/aokamoto/macs2output/${sample}_${stage}_pooled_peaks.narrowPeak /n/capellini_lab/users/aokamoto/Autopod_Project/Human_ATAC/IDR_called_peaks/

echo "Success!"

done
```

## Finally, filter using embryonic brain peak data, generate BED files 

```{bash, eval = FALSE}
for sample in HP1 HP2 HP3 HP4 HP5 HM1 HM2 HM3 HM4 HM5 FP1 FP2 FP3 FP4 FP5 FM1 FM2 FM3 FM4 FM5

do

bedtools subtract -A -a /Users/alexanderokamoto/Desktop/Capellini_Lab/Human_ATAC/Human_Autopod_IDR_peaks/${sample}_late_pooled_peaks_replicate_consolidated_IDR_threshold0.05.narrowPeak -b /Users/alexanderokamoto/Desktop/Capellini_Lab/Brain_ATAC_Files/human_brain_E54_peaks_hg38.bed > /Users/alexanderokamoto/Desktop/Capellini_Lab/Human_ATAC/Human_Autopod_IDR_peaks/Brain_Filtered/${sample}_late_IDR_0.05_brain_filtered.narrowPeak

done

#now create BED files 
for sample in HP1 HP2 HP3 HP4 HP5 HM1 HM2 HM3 HM4 HM5 FP1 FP2 FP3 FP4 FP5 FM1 FM2 FM3 FM4 FM5

do

cut -f 1-3 /Users/alexanderokamoto/Desktop/Capellini_Lab/Human_ATAC/Human_Autopod_IDR_peaks/Brain_Filtered/${sample}_late_IDR_0.05_brain_filtered.narrowPeak > /Users/alexanderokamoto/Desktop/Capellini_Lab/Human_ATAC/Human_Autopod_IDR_peaks/Brain_Filtered/${sample}_late_IDR_0.05_brain_filtered.bed

done
```

##for generating peakCounts
## uses merged IDR peaks file: /human_all_merge_IDR_0.05_brain_filtered.bed

```{bash}}

/n/capellini_lab/users/aokamoto/Autopod_Project/Human_ATAC/IDR_called_peaks/human_all_merge_IDR_0.05_brain_filtered.bed 

#reformat bed
awk 'BEGIN {OFS="\t"} {print $1, $2, $3, $1":"$2"-"$3}' /n/capellini_lab/users/aokamoto/Autopod_Project/Human_ATAC/IDR_called_peaks/human_all_merge_IDR_0.05_brain_filtered.bed > /n/capellini_lab/users/aokamoto/Autopod_Project/Human_ATAC/IDR_called_peaks/human_all_merge_IDR_0.05_brain_filtered_2.bed 

#generate saf file 
awk 'BEGIN{FS=OFS="\t"; print "GeneID\tChr\tStart\tEnd\tStrand"}{print $4, $1, $2+1, $3, "."}' /n/capellini_lab/users/aokamoto/Autopod_Project/Human_ATAC/IDR_called_peaks/human_all_merge_IDR_0.05_brain_filtered_2.bed  > /n/capellini_lab/users/aokamoto/Autopod_Project/Human_ATAC/IDR_called_peaks/human_all_merge_IDR_0.05_brain_filtered.saf

#get feature counts
/n/capellini_lab/users/aokamoto/Software/subread-2.0.2-Linux-x86_64/bin/featureCounts -a /n/capellini_lab/users/aokamoto/Autopod_Project/Human_ATAC/IDR_called_peaks/human_all_merge_IDR_0.05_brain_filtered.saf -F SAF -T 20 -t bam -p -o human_ATAC_peaks_countMatrix.txt $SCRATCH/capellini_lab/Lab/aokamoto/Human_noMT/*.bam
28723_HP2.noMT.bam

#save bam files to more permanent directory
cp $SCRATCH/capellini_lab/Lab/aokamoto/Human_noMT/*.bam /n/capellini_lab/users/aokamoto/Autopod_Project/Human_ATAC
```


## MOUSE

Mouse ATAC-seq sequencing data was processed similarly, except using the mouse mm10 genome and Mac3 software due an update to the software ocurring during the course of this research. The script is below:

```{bash, eval = FALSE}

#load modules
module load python/3.12.5-fasrc01
module load gcc/14.2.0-fasrc01
conda activate bedfiles

#if needed, make directory for output reads

if [ -d "$SCRATCH/capellini_lab/Lab/mouse_cat_ATAC_files" ]; 
then
  echo "Output directory $SCRATCH/capellini_lab/Lab/mouse_cat_ATAC_files exists."
  
else
  mkdir $SCRATCH/capellini_lab/Lab/mouse_cat_ATAC_files
fi

#if needed, make directory for trimmed reads

if [ -d "$SCRATCH/capellini_lab/Lab/Mouse_Trimmed_Reads" ]; 
then
  echo "Output directory $SCRATCH/capellini_lab/Lab/Mouse_Trimmed_Reads exists."
  
else
  mkdir $SCRATCH/capellini_lab/Lab/Mouse_Trimmed_Reads
fi

#if needed, make directory for aligned reads
if [ -d "$SCRATCH/capellini_lab/Lab/Mouse_Aligned" ]; 
then
  echo "Output directory $SCRATCH/capellini_lab/Lab/Mouse_Aligned exists."
  
else
  mkdir $SCRATCH/capellini_lab/Lab/Mouse_Aligned
fi

#if needed, make directory for Picard output
if [ -d "$SCRATCH/capellini_lab/Lab/Mouse_Picard_output" ]; 
then
  echo "Output directory $SCRATCH/capellini_lab/Lab/Mouse_Picard_output/ exists."
  
else
  mkdir $SCRATCH/capellini_lab/Lab/Mouse_Picard_output
fi

#if needed, make directory for Gaspar remove mitochondria output
if [ -d "$SCRATCH/capellini_lab/Lab/Mouse_noMT" ]; 
then
  echo "Output directory $SCRATCH/capellini_lab/Lab/Mouse_noMT exists."
  
else
  mkdir $SCRATCH/capellini_lab/Lab/Mouse_noMT
fi

#concatenate reads if needed
rep_id="rep3"
rep_id_two="Mouse3"

for sample in FM1 FM3 FM5 FP1 FP3 FP5 HM1 HM3 HM5 HP1 HP3 HP5

do 

file1=$(find /n/capellini_lab/users/aokamoto/Autopod_Project/Sequencing_Data/221222_A00794_0761_AHHTVMDSX5_SUB13574/fastq/Mouse3andH29067 -name ${rep_id_two}_${sample}*_R1_001.fastq.gz)
file3=$(find /n/capellini_lab/users/aokamoto/Autopod_Project/Sequencing_Data/221222_A00794_0761_AHHTVMDSX5_SUB13574/fastq/Mouse3andH29067 -name ${rep_id_two}_${sample}*_R2_001.fastq.gz)
file2=$(find /n/capellini_lab/users/aokamoto/Autopod_Project/Sequencing_Data/221129_A00794_0742_AHGWKNDSX5_SUB13550/fastq/H28829_mouse2 -name ${rep_id}_${sample}*_R1_001.fastq.gz)
file4=$(find /n/capellini_lab/users/aokamoto/Autopod_Project/Sequencing_Data/221129_A00794_0742_AHGWKNDSX5_SUB13550/fastq/H28829_mouse2 -name ${rep_id}_${sample}*_R2_001.fastq.gz)

if [[ -f ${file1} && ! -f ${file2} ]]; 
then
  cp ${file1} $SCRATCH/capellini_lab/Lab/mouse_cat_ATAC_files/${rep_id}_${sample}_R1.fastq.gz
  cp ${file3} $SCRATCH/capellini_lab/Lab/mouse_cat_ATAC_files/${rep_id}_${sample}_R2.fastq.gz
  echo "Only ${file1} found."
elif [[ ! -f ${file1} && -f ${file2} ]]; then
  cp ${file2} $SCRATCH/capellini_lab/Lab/mouse_cat_ATAC_files/${rep_id}_${sample}_R1.fastq.gz
  cp ${file4} $SCRATCH/capellini_lab/Lab/mouse_cat_ATAC_files/${rep_id}_${sample}_R2.fastq.gz
  echo "Only ${file2} found."
elif [[ -f ${file1} && -f ${file2} ]]; then
  cat ${file1} ${file2} > $SCRATCH/capellini_lab/Lab/mouse_cat_ATAC_files/${rep_id}_${sample}_R1.fastq.gz
  cat ${file3} ${file4} > $SCRATCH/capellini_lab/Lab/mouse_cat_ATAC_files/${rep_id}_${sample}_R2.fastq.gz
  echo "${file1} and ${file2} found."
fi

/n/capellini_lab/users/aokamoto/Software/NGmerge-master/NGmerge -a -1 $SCRATCH/capellini_lab/Lab/mouse_cat_ATAC_files/${rep_id}_${sample}_R1.fastq.gz -2 $SCRATCH/capellini_lab/Lab/mouse_cat_ATAC_files/${rep_id}_${sample}_R2.fastq.gz -o $SCRATCH/capellini_lab/Lab/Mouse_Trimmed_Reads/${rep_id}_${sample}_trimmed -v -e 30 -n 48

echo "Aligning ${sample} to UCSC mm10 genome."

bowtie2 \
-X 2000 \
-t \
-p 48 \
-x /n/capellini_lab/users/aokamoto/Autopod_Project/MouseReference/Mus_musculus/UCSC/mm10/Sequence/Bowtie2Index/genome \
--very-sensitive \
-1 $SCRATCH/capellini_lab/Lab/Mouse_Trimmed_Reads/${rep_id}_${sample}_trimmed_1.fastq.gz \
-2 $SCRATCH/capellini_lab/Lab/Mouse_Trimmed_Reads/${rep_id}_${sample}_trimmed_2.fastq.gz | samtools view -bS - | samtools sort -o $SCRATCH/capellini_lab/Lab/Mouse_Aligned/${rep_id}_${sample}_coordSort.bam

echo "fixing samtools format."
samtools addreplacerg -r "@RG\tID:RG1\tSM:SampleName\tPL:Illumina\tLB:Library.fa" -o $SCRATCH/capellini_lab/Lab/Mouse_Aligned/${rep_id}_${sample}_coordSort2.bam $SCRATCH/capellini_lab/Lab/Mouse_Aligned/${rep_id}_${sample}_coordSort.bam

echo "Align and sort for ${rep_id} ${sample} completed."

samtools index $SCRATCH/capellini_lab/Lab/Mouse_Aligned/${rep_id}_${sample}_coordSort2.bam $SCRATCH/capellini_lab/Lab/Mouse_Aligned/${rep_id}_${sample}_coordSort.bai

echo "Indexing for ${rep_id} ${sample} completed."

sample_file="$SCRATCH/capellini_lab/Lab/Mouse_Aligned/${rep_id}_${sample}_coordSort2.bam"
outfile="$SCRATCH/capellini_lab/Lab/Mouse_Picard_output/${rep_id}_${sample}_picard-mrkdup.bam"
metrics="$SCRATCH/capellini_lab/Lab/Mouse_Picard_output/${rep_id}_${sample}_picard-mrkdup_metrics.txt"

echo "Marking duplicates for ${rep_id} ${sample}"

java -jar /n/capellini_lab/users/aokamoto/Software/picard-master/picard.jar MarkDuplicates \
      -I $sample_file \
      -O $outfile \
      -M $metrics

#index new file with duplicates removed
samtools index $outfile

inBAM="$SCRATCH/capellini_lab/Lab/Mouse_Picard_output/${rep_id}_${sample}_picard-mrkdup.bam"
noMT="$SCRATCH/capellini_lab/Lab/Mouse_noMT/${rep_id}_${sample}.noMT.bam"

echo "Removing mito reads from ${inBAM} using removeChrom.py"

#output new bam file without mitochondrial reads
samtools view -h $inBAM | python /n/capellini_lab/users/aokamoto/Software/removeChrom.py - - chrM | samtools view -b - > $noMT

echo "Calling peaks for ${sample} using MACS2"

/n/home04/aokamoto/.local/bin/macs3 callpeak \
   -t ${noMT} \
   -n ${rep_id}_${sample} \
   -f BAMPE \
   -g mm \
   --outdir "$SCRATCH/capellini_lab/Lab/macs2output" \
   --nolambda \
   --keep-dup all \
   --bdg \
   --verbose 3

echo "Success against all odds!"
done
```

#Now, call IDR script

```{bash, eval = FALSE}
module load python/3.12.5-fasrc01
module load gcc/14.2.0-fasrc01
conda activate bedfiles

id_1='rep1'
id_2='rep2'
id_3='rep3'

for sample in HP1 HP3 HP5 HM1 HM3 HM5 FP1 FP3 FP5 FM1 FM3 FM5

do

inbam1="$SCRATCH/capellini_lab/Lab/Mouse_noMT/${id_1}_${sample}.noMT.bam"
inbam2="$SCRATCH/capellini_lab/Lab/Mouse_noMT/${id_2}_${sample}.noMT.bam"
inbam3="$SCRATCH/capellini_lab/Lab/Mouse_noMT/${id_3}_${sample}.noMT.bam"

inpeaks1="$SCRATCH/capellini_lab/Lab/macs2output/${id_1}_${sample}_peaks.narrowPeak"
inpeaks2="$SCRATCH/capellini_lab/Lab/macs2output/${id_2}_${sample}_peaks.narrowPeak"
inpeaks3="$SCRATCH/capellini_lab/Lab/macs2output/${id_3}_${sample}_peaks.narrowPeak"

echo "Creating combined peak file for ${sample}"
rm mouse_${sample}_peakset.txt
echo ${inpeaks1} > mouse_${sample}_peakset.txt
echo ${inpeaks2} >> mouse_${sample}_peakset.txt
echo ${inpeaks3} >> mouse_${sample}_peakset.txt

echo "Merging replicates for  ${sample} using samtools"

samtools merge "$SCRATCH/capellini_lab/Lab/Mouse_noMT/${sample}_pooled.bam" ${inbam1} ${inbam2} ${inbam3}

echo "Calling peaks for ${sample} using MACS3"

macs3 callpeak \
-t "$SCRATCH/capellini_lab/Lab/Mouse_noMT/${sample}_pooled.bam" \
-n mouse_${sample}_pooled \
-f BAMPE \
-g mm \
--outdir "$SCRATCH/capellini_lab/Lab/macs2output" \
--nolambda \
--keep-dup all \
--bdg \
--verbose 3
   
echo "Calculating IDR peak set for ${stage} ${sample}"

Rscript IDR_script_APRIL2021_CUSTOM_CUTOFFS.R idr_calling_updated mouse_${sample}_peakset.txt $SCRATCH/capellini_lab/Lab/macs2output/mouse_${sample}_pooled_peaks.narrowPeak /n/capellini_lab/users/aokamoto/Autopod_Project/Mouse_ATAC/IDR_called_peaks/

echo "Success!"

done
```

#brain filtering and bed file generation

```{bash, eval = FALSE}
for sample in HP1 HP3 HP5 HM1 HM3 HM5 FP1 FP3 FP5 FM1 FM3 FM5

do

bedtools subtract -A -a mouse_${sample}_pooled_peaks_replicate_consolidated_IDR_threshold0.05.narrowPeak -b /Users/alexanderokamoto/Desktop/Capellini_Lab/Capellini_lab_Peak_Sets/Brain_ATAC_Files/mouse_all_files_aggregated_brain.bed > Brain_Filtered/mouse_${sample}_IDR_0.05_brain_filtered.narrowPeak

done

#now create BED files 
for sample in HP1 HP3 HP5 HM1 HM3 HM5 FP1 FP3 FP5 FM1 FM3 FM5

do

cut -f 1-4 Brain_Filtered/mouse_${sample}_IDR_0.05_brain_filtered.narrowPeak > Brain_Filtered/mouse_${sample}_IDR_0.05_brain_filtered.bed

done
```

##for generating peakCounts

```{bash}
#create merged bed file
cat mouse_HM1_IDR_0.05_brain_filtered_sorted_mm10.bed mouse_FM1_IDR_0.05_brain_filtered_sorted_mm10.bed	 mouse_FP1_IDR_0.05_brain_filtered_sorted_mm10.bed	 mouse_HM3_IDR_0.05_brain_filtered_sorted_mm10.bed mouse_HM5_IDR_0.05_brain_filtered_sorted_mm10.bed mouse_FM3_IDR_0.05_brain_filtered_sorted_mm10.bed		mouse_FP3_IDR_0.05_brain_filtered_sorted_mm10.bed		mouse_HP1_IDR_0.05_brain_filtered_sorted_mm10.bed mouse_HP3_IDR_0.05_brain_filtered_sorted_mm10.bed mouse_FM5_IDR_0.05_brain_filtered_sorted_mm10.bed		mouse_FP5_IDR_0.05_brain_filtered_sorted_mm10.bed		mouse_HP5_IDR_0.05_brain_filtered_sorted_mm10.bed | sort -k 1,1 -k2,2n | bedtools merge -i stdin > mouse_all_merged_IDR_0.05_brain_filtered_sorted_mm10.bed

#reformat bed file
awk 'BEGIN {OFS="\t"} {print $0, $1":"$2"-"$3}' mouse_all_merged_IDR_0.05_brain_filtered_sorted_mm10.bed > mouse_all_merged_IDR_0.05_brain_filtered_sorted_mm10_2.bed

#create saf file
awk 'BEGIN{FS=OFS="\t"; print "GeneID\tChr\tStart\tEnd\tStrand"}{print $4, $1, $2+1, $3, "."}' mouse_all_merged_IDR_0.05_brain_filtered_sorted_mm10_2.bed > mouse_all_merged_IDR_0.05_brain_filtered_sorted_mm10.saf

#get feature counts  
/n/capellini_lab/users/aokamoto/Software/subread-2.0.2-Linux-x86_64/bin/featureCounts -a /n/capellini_lab/users/aokamoto/Autopod_Project/Mouse_ATAC/IDR_called_peaks/mouse_all_merged_IDR_0.05_brain_filtered_sorted_mm10.saf -F SAF -T 20 -t bam -p -o mouse_ATAC_peaks_countMatrix.txt $SCRATCH/capellini_lab/Lab/Mouse_noMT/*.bam

#save results to a more permanent direcotry
cp $SCRATCH/capellini_lab/Lab/Mouse_noMT/* /n/capellini_lab/users/aokamoto/Autopod_Project/Mouse_ATAC
```

### for generating bigwig files (skeleton script)

```{bash}

#load samtools and deeptools
module load python/3.12.11-fasrc01
source activate base
conda activate /n/home04/aokamoto/.conda/envs/deeptools

#iterate over bams
for bam_file in *noMT.bam; do

 # Extract the base name (e.g., "sample" from "sample.bam")
        base="${bam_file%.bam}"
        echo "Creating bigwig file for $bam_file"
# sort bam files 
        samtools sort -@ 12 $bam_file -o "${base}.sorted.bam"
#index bam files
        samtools index -@ 11 "${base}.sorted.bam"
#index convert bam to bigwig
        bamCoverage -p 12 -b "${base}.sorted.bam" -o "${base}.sorted.bw"
done

```

### for calculating Pearson correlation coefficents between biological replicates (skeleton script)
```{bash}
#load deeptools
module load python/3.12.11-fasrc01
source activate base
conda activate /n/home04/aokamoto/.conda/envs/deeptools

#iterate over tissues each tissue
for sample in HP1 HP2 HP3 HP4 HP5 HM1 HM2 HM3 HM4 HM5 FP1 FP2 FP3 FP4 FP5 FM1 FM2 FM3 FM4 FM5
        do
        multiBigwigSummary bins -b 28597_${sample}.noMT.sorted.bw 28824_${sample}.noMT.sorted.bw 28829_${sample}.noMT.sorted.bw -o ${sample}_early_results.npz
        plotCorrelation -in ${sample}_early_results.npz -c pearson -p heatmap -o plot.png --outFileCorMatrix ${sample}_early_PearsonCorr_bigwigScores.tab
done

```

### for geting TSS enrichment for ATAC-samples
```{r}
## load packages
library(ATACseqQC)
library(ChIPpeakAnno)
library("GenomicAlignments")
library(TxDb.Hsapiens.UCSC.hg38.knownGene) #for human
library(TxDb.Mmusculus.UCSC.mm10.knownGene) #for mouse

library(Rsamtools)

output_file <- "human_TSS"
reps = c("29042", "29067")

txs <- transcripts(TxDb.Hsapiens.UCSC.hg38.knownGene)

#set parameters for reading BAMs
flag <- scanBamFlag(
  isUnmappedQuery      = FALSE,
  isSecondaryAlignment = FALSE,
  isDuplicate          = FALSE
)

param <- ScanBamParam(
  flag  = flag
)

samples = c("FM1", "FM2", "FM3", "FM4", "FM5", "FP1", "FP2", "FP3", "FP4", "FP5", "HM1", "HM2", "HM3", "HM4", "HM5", "HP1", "HP2", "HP3", "HP4", "HP5")

#create df for TSS enrichment results
TSS_df <- data.frame(sample = paste(rep(reps, each = 5), "_", samples, ".noMT.sorted.bam", sep = ""), TSS_score = NA)

for (i in 1:nrow(TSS_df)){
  gal <- readGAlignments(TSS_df$sample[i], param = param)
  tsse <- TSSEscore(gal, txs)
  TSS_df$TSS_score[i] <- tsse$TSSEscore
  write.table(x = TSS_df, file = paste(output_file, ".tsv", sep =""), sep = "\t", )
}
```



